---
title: "Where New Zealand's sheep went"
subtitle: "A regional breakdown of the flock decline, 2002–2025"
author: "Feng Jiang"
date: today
---

```{r setup}
#| include: false
source("R/load.R")
source("R/checks.R")
suppressPackageStartupMessages({
  library(ggplot2); library(scales); library(tidyr); library(knitr)
})

dir.create("outputs/figures", recursive = TRUE, showWarnings = FALSE)

d       <- livestock
regions <- regions_only(d)

fmt <- function(x, digits = 0) format(round(x, digits), big.mark = ",", trim = TRUE)

# Sheep, first year to last year, by region -------------------------------
sheep_change <- regions |>
  filter(livestock_class == "Sheep", year %in% c(2002, 2025)) |>
  select(region, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002, pct = 100 * change / y2002) |>
  arrange(change)

measurable  <- sum(!is.na(sheep_change$change))
decline_sum <- sum(sheep_change$change[sheep_change$change < 0], na.rm = TRUE)
top3        <- head(sheep_change, 3)
top3_share  <- 100 * sum(top3$change) / decline_sum

nat_sheep <- national(d) |> filter(livestock_class == "Sheep")
nat_2002  <- nat_sheep$head[nat_sheep$year == 2002]
nat_2025  <- nat_sheep$head[nat_sheep$year == 2025]
nat_drop  <- nat_2002 - nat_2025
nat_pct   <- 100 * nat_drop / nat_2002

nat_wide <- national(d) |>
  filter(year %in% c(2002, 2025)) |>
  select(livestock_class, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002)

dairy_gain <- nat_wide$change[nat_wide$livestock_class == "Dairy cattle"]
beef_gain  <- nat_wide$change[nat_wide$livestock_class == "Beef cattle"]

# Substitution, all three classes ----------------------------------------
substitution <- regions |>
  filter(year %in% c(2002, 2025)) |>
  select(region, livestock_class, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002) |>
  select(region, livestock_class, change) |>
  pivot_wider(names_from = livestock_class, values_from = change)

top5_regions <- head(sheep_change$region, 5)

dairy_both    <- substitution |> filter(!is.na(Sheep), !is.na(`Dairy cattle`))
n_dairy_falls <- sum(dairy_both$`Dairy cattle` < 0)
n_dairy_both  <- nrow(dairy_both)

# Suppression rate --------------------------------------------------------
supp <- regions |>
  group_by(year, is_census_year) |>
  summarise(cells = n(), suppressed_cells = sum(suppressed), .groups = "drop") |>
  mutate(rate = 100 * suppressed_cells / cells)

# The two flags mean different things and are counted separately. "C" is the
# confidentiality flag Stats NZ used before 2017; "S" is quality suppression.
supp_by_code <- regions |>
  filter(suppressed) |>
  count(year, suppression_code) |>
  left_join(select(supp, year, cells), by = "year") |>
  mutate(
    rate  = 100 * n / cells,
    label = ifelse(suppression_code == "c",
                   "C - confidentiality (pre-2017 method)",
                   "S - quality suppression")
  )

c_years <- sort(unique(supp_by_code$year[supp_by_code$suppression_code == "c"]))
s_years <- sort(unique(supp_by_code$year[supp_by_code$suppression_code == "s"]))

# The same check across every livestock code in the raw extract, not just the
# three used here, so the claim about the C flag is tested on the whole file.
all_c_years <- readr::read_csv(RAW_CSV, col_types = readr::cols(.default = "c")) |>
  janitor::clean_names() |>
  filter(obs_status == "c") |>
  pull(year_agr_agr_003) |> as.integer() |> unique() |> sort()
```

::: {.callout-note appearance="simple"}
Source data: Stats NZ, *Agricultural production statistics*, licensed by Stats NZ
for re-use under the Creative Commons Attribution 4.0 International licence.
:::

## The short version

New Zealand's sheep flock fell by **`r fmt(nat_drop/1e6, 2)` million** between June
2002 and June 2025, a drop of `r fmt(nat_pct, 0)` percent.

Three points, each with one number.

- Three regions — `r paste(top3$region, collapse = ", ")` — account for
  **`r fmt(top3_share, 0)` percent** of the total fall, across the
  `r measurable` regions where both endpoints are published.
- Over the same 23 years the national dairy herd grew by only
  **`r fmt(dairy_gain/1e3, 0)` thousand** cattle, while beef cattle *fell* by
  `r fmt(-beef_gain/1e3, 0)` thousand.
- In `r n_dairy_falls` of the `r n_dairy_both` regions where both series are
  published, dairy cattle fell **alongside** sheep rather than in place of them.

### What this means

The common story is that sheep farms became dairy farms. At a national level the
arithmetic does not support that story. Sheep numbers fell by roughly 16 million
head. Dairy cattle rose by roughly six hundred thousand. Those are different
orders of magnitude.

The picture changes region by region. Canterbury and Southland did add dairy
cattle on a scale worth noticing. Most other regions that lost sheep did not, and
several lost dairy cattle as well. Whatever replaced the sheep, in most of the
country it was not another animal counted in this table.

### What these numbers do not tell you

They are head counts. They say nothing about land area, feed demand, grazing
pressure, farm profitability, or what the land is used for now. A dairy cow and a
sheep are not exchangeable units, and this page does not treat them as such.

## The regional breakdown

```{r}
#| label: fig-contribution
#| fig-cap: "Contribution of each region to the national fall in sheep numbers, June 2002 to June 2025."
plot_dat <- sheep_change |>
  filter(!is.na(change)) |>
  mutate(region = factor(region, levels = rev(region)),
         millions = -change / 1e6)

p1 <- ggplot(plot_dat, aes(x = millions, y = region)) +
  geom_col(fill = "#2d6a4f", width = 0.72) +
  geom_text(aes(label = sprintf("%.2f", millions)),
            hjust = -0.15, size = 3.1, colour = "grey25") +
  annotate("label", x = max(plot_dat$millions) * 0.45,
           y = 4.2,
           label = sprintf("Canterbury, Southland and Otago\naccount for %.0f%% of the fall",
                           top3_share),
           hjust = 0, size = 3.4, label.size = 0, fill = "grey96", colour = "grey20") +
  scale_x_continuous(labels = label_comma(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.14))) +
  labs(
    title    = "Three South Island regions account for over half the national fall",
    subtitle = "Reduction in sheep numbers, millions of head, June 2002 to June 2025",
    x = "Million head fewer", y = NULL,
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title.position = "plot",
        plot.title    = element_text(face = "bold"),
        plot.caption  = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/01-regional-contribution.png", p1,
       width = 9, height = 6, dpi = 200, bg = "white")
p1
```

Two regions are missing from that chart. Auckland and Nelson have a suppressed
2025 value, so no change can be computed for them. They are excluded and counted,
never imputed and never set to zero.

## Did dairy replace the sheep?

```{r}
#| label: fig-substitution
#| fig-cap: "Change in sheep, dairy cattle and beef cattle in the five regions that lost the most sheep."
dumb <- substitution |>
  filter(region %in% top5_regions) |>
  pivot_longer(-region, names_to = "livestock_class", values_to = "change") |>
  mutate(
    region = factor(region, levels = rev(top5_regions)),
    livestock_class = factor(livestock_class,
                             levels = c("Sheep", "Dairy cattle", "Beef cattle")),
    thousands = change / 1e3
  )

p2 <- ggplot(dumb, aes(x = thousands, y = region)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_line(aes(group = region), colour = "grey75", linewidth = 0.8) +
  geom_point(aes(colour = livestock_class), size = 3.4) +
  scale_colour_manual(values = c("Sheep" = "#2d6a4f",
                                 "Dairy cattle" = "#c1666b",
                                 "Beef cattle" = "#4a6fa5"), name = NULL) +
  scale_x_continuous(labels = label_comma(), expand = expansion(mult = 0.07)) +
  labs(
    title    = "Where sheep left, cattle mostly did not arrive in comparable numbers",
    subtitle = "Change in animals, thousands of head, June 2002 to June 2025, five largest sheep declines",
    x = "Thousand head, change over the period", y = NULL,
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/02-substitution.png", p2,
       width = 9, height = 5.5, dpi = 200, bg = "white")
p2
```

Stated plainly, per hundred sheep lost:

```{r}
#| label: tbl-substitution
substitution |>
  filter(region %in% top5_regions) |>
  transmute(
    Region = region,
    `Sheep lost` = fmt(-Sheep),
    `Dairy cattle per 100 sheep lost` = sprintf("%+.1f", -100 * `Dairy cattle` / Sheep),
    `Beef cattle per 100 sheep lost`  = sprintf("%+.1f", -100 * `Beef cattle` / Sheep)
  ) |>
  kable(align = "lrrr")
```

Canterbury is the one region where the substitution story carries real weight.
Hawkes Bay and Manawatu-Wanganui lost sheep *and* cattle. A negative figure in
this table means that class of animal declined as well.

## Data quality checks

Five rules run over the analysis table on every render.

```{r}
#| label: tbl-validation
kable(results, align = "lrrrr")
```

The interesting result is not in that table. It is in the reconciliation: the
published national total against the sum of the regions.

```{r}
#| label: tbl-recon
reconciliation |>
  filter(livestock_class == "Sheep") |>
  left_join(select(coverage, year, regions_present, regions_suppressed), by = "year") |>
  transmute(
    Year = year,
    `Regions published` = regions_present,
    `Regions suppressed` = regions_suppressed,
    `Sum of regions` = fmt(region_sum),
    `Published national` = fmt(published),
    `Residual %` = sprintf("%.3f", residual_pct)
  ) |>
  kable(align = "rrrrrr")
```

The residual is zero, to the head, in every year where all seventeen regions are
published and no cell is suppressed. It becomes non-zero for exactly two reasons,
and only those two: a region is suppressed, or a region code is absent from the
table for that year. It is not a join error and it is not perturbation.

### Suppression is not constant over time

```{r}
#| label: fig-suppression
#| fig-cap: "Share of regional cells suppressed, by year."
p3 <- ggplot(supp_by_code, aes(x = year, y = rate)) +
  geom_col(aes(fill = label), width = 0.7) +
  scale_fill_manual(values = c("C - confidentiality (pre-2017 method)" = "#e07a5f",
                               "S - quality suppression" = "#6c8ea4"), name = NULL) +
  scale_x_continuous(breaks = seq(2002, 2025, 3), limits = c(2001, 2026)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(
    title    = "Two different flags, two different stories",
    subtitle = "Percentage of region-by-livestock-class cells withheld: sheep, dairy and beef cattle",
    x = NULL, y = "Cells suppressed",
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/03-suppression-rate.png", p3,
       width = 9, height = 5, dpi = 200, bg = "white")
p3
```

Reading that chart as a single rising trend would be wrong, and the `OBS_STATUS`
column is what makes the difference visible. Two flags are in use, and they are
not the same mechanism.

The `C` flag is confidentiality suppression. Among the three livestock classes
used here it appears in `r paste(c_years, collapse = ", ")` and in no other year.
Checked across every one of the 44 livestock codes in the raw extract, it appears
in `r paste(range(all_c_years), collapse = "–")` and **never after
`r max(all_c_years)`**.

That is exactly what the published methodology predicts. Stats NZ states that
prior to 2017, figures affected by the confidentiality provisions were denoted by
`C`, and that since 2017 confidentiality has instead been implemented by an input
perturbation method that adds noise at individual farm level, so output figures
no longer have to be replaced with `C`. The flag stops in the file in the same
year the method changed.

The `S` flag is quality suppression, applied where sampling errors or imputation
levels are high. It runs from `r min(s_years)` to `r max(s_years)`, peaking at
`r sprintf("%.1f", max(supp$rate))` percent of cells in
`r supp$year[which.max(supp$rate)]`. The census years 2017 and 2022 sit below the
survey years around them, which is what a rule keyed to sampling error would
predict, since a census has no sampling error to exceed a threshold.

So the 2012 spike is not a quality problem at all. All four withheld cells that
year carry `C`, not `S`. Separating the two flags turns an apparent anomaly into
the expected behaviour of a confidentiality rule that was replaced in 2017.

One qualification remains. I cannot tell from this extract whether the absence of
any flag before 2012 means no cell was withheld, or that the flag was not carried
in this export. The chart shows what the file contains.

Two of the affected regions are also explained in the published methodology.
Sampling error could not be calculated for Nelson because only one responding
unit was observed in each sampled stratum, and at least two are needed to
estimate a within-stratum variance. Nelson is one of the two regions this
analysis has to drop.

### Independent recomputation

Stats NZ publishes a national sheep figure for June 2024. Recomputing it from the
regional cells in the same extract gives a different number, and the difference
is fully explained.

```{r}
#| label: tbl-recompute
r24 <- reconciliation |> filter(livestock_class == "Sheep", year == 2024)
data.frame(
  Quantity = c("Published, Total New Zealand, June 2024",
               "Recomputed as the sum of published regions",
               "Difference"),
  Head = c(fmt(r24$published), fmt(r24$region_sum), fmt(r24$residual))
) |> kable(align = "lr")
```

The difference is Auckland, whose 2024 value is suppressed. Filling it with zero
would have produced a total that looked tidier and was wrong by
`r fmt(r24$residual)` head.

## Method

1. **One table, one vintage.** Stats NZ table `AGR_AGR_003`, *Livestock Numbers
   by Regional Council*, dataflow `STATSNZ:AGR_AGR_003(1.0)`, final vintage,
   downloaded once and committed with its SHA-256. Full provenance is in
   [`data-raw/SOURCE.md`](https://github.com/Kenchch/nz-sheep-decline-by-region/blob/main/data-raw/SOURCE.md).
2. **Codes verified before code was written.** The extract carries numeric codes
   and no labels. Sheep, dairy cattle and beef cattle were each confirmed against
   the published June 2024 national figures before any analysis was run. The
   `LIVESTOCK` codelist holds 44 entries that nest inside one another; adding all
   44 for 2024 gives 115.7 million animals against a published 33.8 million. Only
   the three verified codes are used.
3. **Aggregates are separated, not summed.** `AREA` codes 10, 19 and 20 are the
   North Island, South Island and New Zealand totals. They are flagged in the
   analysis table and excluded from every regional sum.
4. **Suppressed cells stay suppressed.** A suppressed cell becomes `NA` and
   carries a flag. It is never filled with zero and never dropped. Regions
   missing either endpoint are excluded from the change calculation and counted
   in the text.
5. **Reconciliation is reported, not asserted.** No fixed tolerance is set
   between the regional sum and the published total. The residual is computed for
   every year and published above.

### Release schedule

This analysis uses the final vintage throughout. Up to and including the 2024
survey, Stats NZ published a provisional release in mid-December and a final
release in mid-May of the following year. From the 2025 survey those two were
replaced by a single final release in mid-April.

### Survey design

The reference period is the year ended 30 June. Censuses were held in 2002, 2007,
2012, 2017 and 2022, with annual sample surveys in the intervening years. The
sample is stratified by regional council area, ANZSIC06 group and size group. The
2025 survey drew 24,600 geographic locations from an estimated eligible
population of 45,800; note that a geographic location is the statistical unit and
is not the same thing as a farm.

The series starts at 2002 because Stats NZ states that the population for the
2002 census and subsequent surveys differs from that of earlier agricultural
censuses and surveys, and that figures from 2002 onward may not be directly
comparable with previous years. That is also why the 1994 observation carried in
the same table is excluded here. There was no agricultural production survey at
all in 1997, 1998 or 2001.

Values for non-respondents are imputed by random hot deck imputation, with the
imputation class formed from regional council area, ANZSIC06 group and production
data from previous years. This matters more than it may appear: for the 2025
survey Stats NZ reports that **30 percent** of the total sheep estimate was
imputed, against a relative sampling error of 3 percent at 95 percent confidence.
The response rate for the 2025 final release was 72.1 percent. The collection is
run by Stats NZ in partnership with the Ministry for Primary Industries.

### What was excluded, and why

- **Territorial authority geography** — the surveys are designed to produce
  results to regional council level and the censuses to territorial authority
  level, so no annual series exists below the region.
- **Horticulture** — not collected in 2004, 2006, 2008, 2010, 2013, 2015, 2016,
  2018, 2021, 2023 or 2025, which is most of this window.
- **Forestry production** — Stats NZ no longer collects or compiles forestry
  production statistics; they are published by MPI. Stats NZ does still collect
  forestry *land use*. The 2025 release also notes that additional National
  Exotic Forest Description data has been received and that revisions to earlier
  land use and forestry statistics, especially for 2016 to 2021, are planned.
- **Pre-2002 data** — the population change described above, plus the absence of
  any survey in 1997, 1998 and 2001.
- **A choropleth map** — a sorted bar chart answers "which regions contributed
  most" and a map does not, because a map encodes land area rather than
  magnitude. Canterbury, Otago and Southland are also three of the largest
  regions by area, which is precisely the confound a map would introduce.
- **A second data set** — the question is answerable inside one table, and every
  extra source adds provenance to defend without adding evidence.

### Implications for output design

Three observations follow from the checks above, and they concern the shape of
the output rather than the sheep.

Regional council is the only geography available annually; territorial authority
exists only in census years, so any sub-regional time series is structurally
unavailable rather than merely unpublished. Suppression now runs above ten
percent of cells and is concentrated in the same small regions — Auckland,
Nelson, Chatham Islands — so those regional trends are already unusable, and a
suppression-aware grouping that pooled them would preserve more signal than
suppressing them one at a time does. And the alternating census-and-survey design
makes every regional series in this table structurally uneven, which belongs on
the output rather than being left for a user to discover from a reconciliation
residual.

There is also a class of question this data cannot answer at all: anything at
farm level, or any breakdown finer than what is published. The access route for
those questions is approved microdata under the Five Safes framework, which I
have not used here.

### Sources for the methodological statements

Every statement about the collection above was read on the page cited here, not
carried over from secondary summaries.

| Statement | Source |
|---|---|
| Census years 2002, 2007, 2012, 2017, 2022; sample surveys in between; eligible population 45,800 and sample 24,600 geographic locations; stratification by regional council, ANZSIC06 group and size group; 2025 response rate 72.1 percent; sampling error and imputation levels by line code; GST registration level of \$60,000 giving "partial and unquantifiable" coverage below it; random hot deck imputation and imputation class; `C` before 2017 and input perturbation since; `S` for data with high sample errors or imputation levels; Nelson variance not estimable; horticulture years; forestry production now MPI; Tier 1 statistics | [Agricultural Production Statistics: June 2025 (Final), DataInfo+](https://datainfoplus.stats.govt.nz/item/nz.govt.stats/a9908d00-4827-46b2-a38a-310c0c75b029) |
| Programme run in partnership with MPI; provisional mid-December plus final mid-May up to 2024, single final mid-April from 2025; 2002 population differs from earlier collections; no survey in 1997, 1998, 2001; regional council annually and territorial authority in censuses | [Agriculture Production Statistics series, DataInfo+](https://datainfoplus.stats.govt.nz/item/nz.govt.stats/6362a469-f374-412e-ac25-d76fd0962003) |
| Release date of 16 April 2026 for the year to June 2025; NEFD data and planned revisions for 2016–2021 | [Agricultural production statistics: Year to June 2025](https://www.stats.govt.nz/information-releases/agricultural-production-statistics-year-to-june-2025/) |
| Mana Ōrite Relationship Agreement signed with the Data Iwi Leaders Group of the National Iwi Chairs Forum on 30 October 2019 | [Mana Ōrite Relationship Agreement](https://www.stats.govt.nz/about-us/what-we-do/mana-orite-relationship-agreement/) |

Statements I could not source on those pages are not made anywhere in this
report. Pages read 4 September 2026.

## Limitations

- Non-census years are **sample estimates**. I have not recomputed their sampling
  errors and I have applied no weighting of my own. A year-on-year movement in a
  single small region should not be read as real change.
- Suppressed cells are retained as `NA` and counted. Two regions are excluded
  from the 2002–2025 change entirely for that reason.
- The sheep-to-cattle comparisons are **ratios of head counts**. They are not
  claims about stocking rate, feed demand, land use or emissions.
- The frame is Stats NZ's Business Register, which is built on GST registration.
  The compulsory registration level is \$60,000, and Stats NZ describes coverage
  of units below that level as "partial and unquantifiable".
- **Imputation is large.** Stats NZ reports 30 percent of the 2025 total sheep
  estimate as imputed. Every regional figure in a survey year carries that, and
  this analysis does nothing to correct for it.
- Since 2017, confidentiality has been implemented by input perturbation, which
  adds noise at individual farm level. It is small relative to the effects
  discussed here, but it is present in every year from 2017 on.
- Choosing 2002 and 2025 as endpoints is a choice. A 2015–2025 window gives the
  same qualitative answer, with the same three regions accounting for 59 percent
  of the fall rather than 54 percent.

## Ethics

This analysis uses published regional aggregates only. No microdata and no IDI
access is involved, so the Five Safes framework does not apply to it.

Stats NZ lists Tatauranga Umanga Māori among the internal users of this
collection, and signed the Mana Ōrite Relationship Agreement with the Data Iwi
Leaders Group of the National Iwi Chairs Forum in October 2019. Any Māori-specific
cut of this data therefore sits inside an established relationship between Stats
NZ and iwi Māori. It does not sit with an unaffiliated analyst working from
published aggregates, which is why this report does not attempt one and does not
characterise the terms of that agreement.

## Reproduce

```bash
quarto render
```

`R/load.R` reads and tidies the pinned extract, `R/checks.R` runs the validation
rules, and `index.qmd` produces this page and writes the three figures into
`outputs/figures/`.

```{r}
#| label: session
#| code-fold: false
sessionInfo()
```
