---
title: "Where New Zealand's sheep went"
subtitle: "A regional breakdown of the flock decline, 2002–2025"
author: "Feng Jiang"
date: today
---

```{r setup}
#| include: false
source("R/checks.R")   # which sources R/load.R, runs the rules and reconciles
suppressPackageStartupMessages({
  library(ggplot2); library(scales); library(tidyr); library(knitr)
})

dir.create("outputs/figures", recursive = TRUE, showWarnings = FALSE)

d       <- livestock
regions <- regions_only(d)

fmt <- function(x, digits = 0) format(round(x, digits), big.mark = ",", trim = TRUE)

# Sheep, first year to last year, by region -------------------------------
sheep_change <- regions |>
  filter(livestock_class == "Sheep", year %in% c(2002, 2025)) |>
  select(region, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002, pct = 100 * change / y2002) |>
  arrange(change)

measurable  <- sum(!is.na(sheep_change$change))
decline_sum <- sum(sheep_change$change[sheep_change$change < 0], na.rm = TRUE)
top3        <- head(sheep_change, 3)
top3_share  <- 100 * sum(top3$change) / decline_sum

nat_sheep <- national(d) |> filter(livestock_class == "Sheep")
nat_2002  <- nat_sheep$head[nat_sheep$year == 2002]
nat_2025  <- nat_sheep$head[nat_sheep$year == 2025]
nat_drop  <- nat_2002 - nat_2025
nat_pct   <- 100 * nat_drop / nat_2002

nat_wide <- national(d) |>
  filter(year %in% c(2002, 2025)) |>
  select(livestock_class, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002)

dairy_gain <- nat_wide$change[nat_wide$livestock_class == "Dairy cattle"]
beef_gain  <- nat_wide$change[nat_wide$livestock_class == "Beef cattle"]

# Substitution, all three classes ----------------------------------------
substitution <- regions |>
  filter(year %in% c(2002, 2025)) |>
  select(region, livestock_class, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002) |>
  select(region, livestock_class, change) |>
  pivot_wider(names_from = livestock_class, values_from = change)

top5_regions <- head(sheep_change$region, 5)

dairy_both    <- substitution |> filter(!is.na(Sheep), !is.na(`Dairy cattle`))
n_dairy_falls <- sum(dairy_both$`Dairy cattle` < 0)
n_dairy_both  <- nrow(dairy_both)

# Regions among the top five that lost cattle as well as sheep, named from the
# data rather than typed into the prose.
lost_both <- substitution |>
  filter(region %in% top5_regions, `Dairy cattle` < 0 | `Beef cattle` < 0) |>
  pull(region) |> sort() |> paste(collapse = " and ")

# Endpoint sensitivity ----------------------------------------------------
# The same headline statistic recomputed from every available start year.
# Asserted in prose, this claim was wrong. Computed, it is a better finding.
top3_for <- function(y0, y1 = 2025) {
  ch <- regions |>
    filter(livestock_class == "Sheep", year %in% c(y0, y1)) |>
    select(region, year, head) |>
    pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
    mutate(change = .data[[paste0("y", y1)]] - .data[[paste0("y", y0)]]) |>
    filter(!is.na(change)) |>
    arrange(change)
  dec <- sum(ch$change[ch$change < 0])
  data.frame(
    start   = y0,
    end     = y1,
    regions = nrow(ch),
    fall    = -dec,
    share   = 100 * sum(head(ch$change, 3)) / dec,
    top3    = paste(head(ch$region, 3), collapse = ", ")
  )
}

sensitivity <- do.call(rbind, lapply(2002:2020, top3_for))

# Split on the separator rather than substring-matching: no region name is a
# substring of another today, but a match that silently over-counts if that
# ever changed is not the kind of code this repository argues for.
in_top3 <- function(name) {
  sum(vapply(strsplit(sensitivity$top3, ", ", fixed = TRUE),
             function(x) name %in% x, logical(1)))
}

# Design-matched comparison: census start, census end.
matched <- top3_for(2002, 2022)

# Suppression rate --------------------------------------------------------
supp <- regions |>
  group_by(year, is_census_year) |>
  summarise(cells = n(), suppressed_cells = sum(suppressed), .groups = "drop") |>
  mutate(rate = 100 * suppressed_cells / cells)

# The two flags mean different things and are counted separately. "C" is the
# confidentiality flag Stats NZ used before 2017; "S" is quality suppression.
supp_by_code <- regions |>
  filter(suppressed) |>
  count(year, suppression_code) |>
  left_join(select(supp, year, cells), by = "year") |>
  mutate(
    rate  = 100 * n / cells,
    label = ifelse(suppression_code == "c",
                   "C - confidentiality (pre-2017 method)",
                   "S - quality suppression")
  )

c_years <- sort(unique(supp_by_code$year[supp_by_code$suppression_code == "c"]))
s_years <- sort(unique(supp_by_code$year[supp_by_code$suppression_code == "s"]))

# The same check across every livestock code in the raw extract, not just the
# three used here, so the claim about the C flag is tested on the whole file.
all_c_years <- all_suppression_years("c")
```

::: {.callout-note appearance="simple"}
Source data: Stats NZ, *Agricultural production statistics*, licensed by Stats NZ
for re-use under the Creative Commons Attribution 4.0 International licence.
:::

## The short version

New Zealand's sheep flock fell by **`r fmt(nat_drop/1e6, 2)` million** between June
2002 and June 2025, a drop of `r fmt(nat_pct, 0)` percent.

Three points, each with one number.

- Three regions — `r paste(top3$region, collapse = ", ")` — account for
  **`r fmt(top3_share, 0)` percent** of the total fall, across the
  `r measurable` regions where both endpoints are published.
- Over the same 23 years the national dairy herd grew by only
  **`r fmt(dairy_gain/1e3, 0)` thousand** cattle, while beef cattle *fell* by
  `r fmt(-beef_gain/1e3, 0)` thousand.
- In `r n_dairy_falls` of the `r n_dairy_both` regions where both series are
  published, dairy cattle fell **alongside** sheep rather than in place of them.

### What this means

The common story is that sheep farms became dairy farms. At a national level the
arithmetic does not support that story. Sheep numbers fell by roughly 16 million
head. Dairy cattle rose by roughly six hundred thousand. Those are different
orders of magnitude.

The picture changes region by region. Canterbury and Southland did add dairy
cattle on a scale worth noticing. Most other regions that lost sheep did not, and
several lost dairy cattle as well. Whatever replaced the sheep, in most of the
country it was not another animal counted in this table.

### What these numbers do not tell you

They are head counts. They say nothing about land area, feed demand, grazing
pressure, farm profitability, or what the land is used for now. A dairy cow and a
sheep are not exchangeable units, and this page does not treat them as such.

## The regional breakdown

```{r}
#| label: fig-contribution
#| fig-cap: "Contribution of each region to the national fall in sheep numbers, June 2002 to June 2025."
#| fig-alt: "Horizontal bar chart of 15 regions ordered by how many sheep they
#|   lost between 2002 and 2025. Canterbury is largest at 3.57 million head
#|   fewer, then Southland at 2.91 million and Otago at 2.27 million; together
#|   these three make up just over half the national fall. The remaining
#|   regions range from Manawatū-Whanganui at 2.25 million down to the Chatham
#|   Islands at 0.01 million."
plot_dat <- sheep_change |>
  filter(!is.na(change)) |>
  mutate(region = factor(region, levels = rev(region)),
         millions = -change / 1e6)

p1 <- ggplot(plot_dat, aes(x = millions, y = region)) +
  geom_col(fill = "#2d6a4f", width = 0.72) +
  geom_text(aes(label = sprintf("%.2f", millions)),
            hjust = -0.15, size = 3.1, colour = "grey25") +
  annotate("label", x = max(plot_dat$millions) * 0.45,
           y = 4.2,
           label = sprintf("%s\naccount for %.0f%% of the fall",
                           paste(top3$region, collapse = ", "), top3_share),
           hjust = 0, size = 3.4, label.size = 0, fill = "grey96", colour = "grey20") +
  scale_x_continuous(labels = label_comma(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.14))) +
  labs(
    title    = "Three South Island regions account for over half the national fall",
    subtitle = "Reduction in sheep numbers, millions of head, June 2002 to June 2025",
    x = "Million head fewer", y = NULL,
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title.position = "plot",
        plot.title    = element_text(face = "bold"),
        plot.caption  = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/01-regional-contribution.png", p1,
       width = 9, height = 6, dpi = 200, bg = "white")
p1
```

Two regions are missing from that chart. Auckland and Nelson have a suppressed
2025 value, so no change can be computed for them. They are excluded and counted,
never imputed and never set to zero.

## Did dairy replace the sheep?

```{r}
#| label: fig-substitution
#| fig-cap: "Change in sheep, dairy cattle and beef cattle in the five regions that lost the most sheep."
#| fig-alt: "Dot chart for the five regions with the largest sheep declines.
#|   In each region the sheep marker sits far to the left of zero, between 1.3
#|   and 3.6 million head fewer, while the dairy and beef cattle markers sit
#|   close to zero. Canterbury has the largest dairy gain at about 795 thousand
#|   and Southland about 281 thousand; in Hawke's Bay and Manawatū-Whanganui
#|   both cattle markers fall below zero, meaning cattle declined as well."
dumb <- substitution |>
  filter(region %in% top5_regions) |>
  pivot_longer(-region, names_to = "livestock_class", values_to = "change") |>
  mutate(
    region = factor(region, levels = rev(top5_regions)),
    livestock_class = factor(livestock_class,
                             levels = c("Sheep", "Dairy cattle", "Beef cattle")),
    thousands = change / 1e3
  )

p2 <- ggplot(dumb, aes(x = thousands, y = region)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_line(aes(group = region), colour = "grey75", linewidth = 0.8) +
  # Shape as well as colour, so colour is not the only channel carrying the
  # distinction. The green and the red-brown are close under red-green colour
  # vision deficiency; WCAG 1.4.1 asks that colour not be used alone.
  geom_point(aes(colour = livestock_class, shape = livestock_class), size = 3.4) +
  scale_colour_manual(values = c("Sheep" = "#2d6a4f",
                                 "Dairy cattle" = "#c1666b",
                                 "Beef cattle" = "#4a6fa5"), name = NULL) +
  scale_shape_manual(values = c("Sheep" = 16, "Dairy cattle" = 17,
                                "Beef cattle" = 15), name = NULL) +
  scale_x_continuous(labels = label_comma(), expand = expansion(mult = 0.07)) +
  labs(
    title    = "Where sheep left, cattle mostly did not arrive in comparable numbers",
    subtitle = "Change in animals, thousands of head, June 2002 to June 2025, five largest sheep declines",
    x = "Thousand head, change over the period", y = NULL,
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/02-substitution.png", p2,
       width = 9, height = 5.5, dpi = 200, bg = "white")
p2
```

Stated plainly, per hundred sheep lost:

```{r}
#| label: tbl-substitution
substitution |>
  filter(region %in% top5_regions) |>
  transmute(
    Region = region,
    `Sheep lost` = fmt(-Sheep),
    `Dairy cattle per 100 sheep lost` = sprintf("%+.1f", -100 * `Dairy cattle` / Sheep),
    `Beef cattle per 100 sheep lost`  = sprintf("%+.1f", -100 * `Beef cattle` / Sheep)
  ) |>
  kable(align = "lrrr")
```

Canterbury accounts for by far the largest absolute dairy increase of any region,
at `r fmt(substitution$"Dairy cattle"[substitution$region == "Canterbury"]/1e3)`
thousand head. It is not, however, the highest *ratio*: West Coast gained
`r sprintf("%.0f", -100 * substitution$"Dairy cattle"[substitution$region == "West Coast"] / substitution$Sheep[substitution$region == "West Coast"])`
dairy cattle per 100 sheep lost, on a base so small — under
`r fmt(-substitution$Sheep[substitution$region == "West Coast"]/1e3)` thousand
sheep — that the ratio says more about the denominator than about land use.
`r lost_both` lost sheep *and* cattle. A negative figure in
this table means that class of animal declined as well.

## Data quality checks

Five rules run over the analysis table on every render.

```{r}
#| label: tbl-validation
kable(results, align = "lrrrr")
```

Four pass. The fifth fires `r results$fails[results$rule == "census_year_reported"]`
times, and that is a finding rather than a defect. `census_year_reported` asserts
that a census year — a full-coverage collection, with no sampling error to
exceed a threshold — should not have withheld cells. It is wrong. Census years do
withhold cells, because withholding is about confidentiality and imputation
quality, not about coverage. The rule earns its place by being falsified.

### Showing that the rules can fail

An all-green table is not evidence. A reader cannot tell from it whether the
rules are sound or merely too loose to fire. So the same five rules are also run
against a deliberately corrupted copy of the analysis table: one negative count,
one withheld cell filled with a zero, one census-year value removed, one region
label deleted, and one duplicated row.

```{r}
#| label: tbl-validation-corrupted
results |>
  select(rule, `real data` = fails) |>
  left_join(select(results_corrupted, rule, `corrupted copy` = fails),
            by = "rule") |>
  kable(align = "lrr", col.names = c("Rule", "Fails on real data",
                                     "Fails on corrupted copy"))
```

Each corruption moves exactly one rule. `no_duplicate_cells` moves by two
because a duplicate makes both members of the pair non-unique, and
`census_year_reported` moves from `r results$fails[results$rule == "census_year_reported"]`
to `r results_corrupted$fails[results_corrupted$rule == "census_year_reported"]`
because it already fails nine times on the real data. The corrupted rows are
chosen at distinct indices and the function asserts as much, so no single
corruption can be counted twice.

That is what makes the left-hand column mean something: the zeroes are the
result of rules that demonstrably fire when there is something to fire at.

The more interesting result is in the reconciliation: the published national
total against the sum of the regions.

```{r}
#| label: tbl-recon
reconciliation |>
  filter(livestock_class == "Sheep") |>
  left_join(select(coverage, year, regions_present, regions_suppressed), by = "year") |>
  transmute(
    Year = year,
    `Regions published` = regions_present,
    `Regions suppressed` = regions_suppressed,
    `Sum of regions` = fmt(region_sum),
    `Published national` = fmt(published),
    `Residual (head)` = fmt(residual),
    `Residual %` = sprintf("%.3f", residual_pct)
  ) |>
  kable(align = "rrrrrrr")
```

The residual column is reported in head as well as in percent on purpose. At a
base of 30 million, a percentage rounded to three decimal places renders a
discrepancy of one animal as `0.000`, which would let the table appear to
support a stronger claim than the numbers do.

Across all three livestock classes the residual separates into three tiers.

```{r}
#| label: tbl-tiers
residual_tiers |>
  count(tier, name = "Class-years") |>
  left_join(
    residual_tiers |>
      group_by(tier) |>
      summarise(`Largest absolute residual (head)` = fmt(max(abs(residual))),
                .groups = "drop"),
    by = "tier"
  ) |>
  rename(Tier = tier) |>
  kable(align = "lrr")
```

In the `r sum(residual_tiers$tier == "Exact")` class-years where all seventeen
regions are published and nothing is withheld, the regions sum to the published
national total **exactly, to the head**.

Completeness is the criterion for these tiers, not the residual. That matters
here: `r sum(residual_tiers$residual == 0)` class-years reconcile to zero, one
more than the Exact tier contains, because one class-year with an incomplete
region row happens to reconcile exactly anyway.

In `r sum(residual_tiers$tier == "Fully published, off by a few head")` further
fully-published class-years the residual never exceeds
`r max(abs(residual_tiers$residual[residual_tiers$tier == "Fully published, off by a few head"]))`
head, on bases between
`r fmt(min(residual_tiers$published[residual_tiers$tier == "Fully published, off by a few head"])/1e6, 1)`
and
`r fmt(max(residual_tiers$published[residual_tiers$tier == "Fully published, off by a few head"])/1e6, 1)`
million. I cannot attribute that from published data. It
is consistent with the national aggregate being rounded independently of the
regional cells, but I have not been able to source that and do not assert it.
Nor can I rule out input perturbation: only two fully-published class-years fall
after 2017, when perturbation was introduced, which is far too few to say
anything either way.

Every residual larger than that — and the next smallest is
`r fmt(min(abs(residual_tiers$residual[residual_tiers$tier == "Incomplete region row" & residual_tiers$residual != 0])))`
head — belongs to a year whose region row is incomplete, because a region is
withheld or because a region code is absent from that year's table.

What the one-head residuals are *not* is an artefact of this analysis. The same
discrepancy appears between cells that this analysis never sums. Adding the two
published island totals and comparing them against the published national total —
three aggregate cells, always published, never withheld — gives a difference of
one head in
`r sum(island_reconciliation$residual != 0)` of the
`r nrow(island_reconciliation)` class-years, and never more than
`r max(abs(island_reconciliation$residual))`.

```{r}
#| label: tbl-island
island_reconciliation |>
  filter(residual != 0) |>
  transmute(
    Year = year,
    Class = livestock_class,
    `North Island` = fmt(a10),
    `South Island` = fmt(a19),
    `Sum` = fmt(a10 + a19),
    `Published NZ` = fmt(a20),
    `Difference` = fmt(residual)
  ) |>
  kable(align = "rlrrrrr")
```

So the one-head discrepancies are a property of the published table, not of the
join. That does not explain them, but it does locate them.

### Suppression is not constant over time

```{r}
#| label: fig-suppression
#| fig-cap: "Share of regional cells withheld, by year and by flag."
#| fig-alt: "Bar chart of the percentage of regional cells withheld each year
#|   from 2002 to 2025, coloured by flag. Nothing is withheld from 2002 to
#|   2011. A single orange bar in 2012 marks the C confidentiality flag at
#|   about 8 percent. Blue bars for the S quality-suppression flag begin in
#|   2014 and run to 2025, peaking near 18 percent in 2021, with the census
#|   years 2017 and 2022 lower than the survey years around them."
p3 <- ggplot(supp_by_code, aes(x = year, y = rate)) +
  geom_col(aes(fill = label), width = 0.7) +
  scale_fill_manual(values = c("C - confidentiality (pre-2017 method)" = "#e07a5f",
                               "S - quality suppression" = "#6c8ea4"), name = NULL) +
  scale_x_continuous(breaks = seq(2002, 2025, 3), limits = c(2001, 2026)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(
    title    = "Two different flags, two different stories",
    subtitle = "Percentage of region-by-livestock-class cells withheld: sheep, dairy and beef cattle",
    x = NULL, y = "Cells suppressed",
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/03-suppression-rate.png", p3,
       width = 9, height = 5, dpi = 200, bg = "white")
p3
```

Reading that chart as a single rising trend would be wrong, and the `OBS_STATUS`
column is what makes the difference visible. Two flags are in use, and they are
not the same mechanism.

The `C` flag is confidentiality suppression. Among the three livestock classes
used here it appears in `r paste(c_years, collapse = ", ")` and in no other year.
Checked across every one of the 44 livestock codes in the raw extract, it appears
in `r paste(range(all_c_years), collapse = "–")` and **never after
`r max(all_c_years)`**.

That is exactly what the published methodology predicts. Stats NZ states that
prior to 2017, figures affected by the confidentiality provisions were denoted by
`C`, and that since 2017 confidentiality has instead been implemented by an input
perturbation method that adds noise at individual farm level, so output figures
no longer have to be replaced with `C`. The flag stops in the file in the same
year the method changed.

The `S` flag is quality suppression, applied where sampling errors or imputation
levels are high. It runs from `r min(s_years)` to `r max(s_years)`, peaking at
`r sprintf("%.1f", max(supp$rate))` percent of cells in
`r supp$year[which.max(supp$rate)]`. The census years 2017 and 2022 sit below the
survey years around them — but not at zero, and the difference between those two
statements is the mechanism.

The published rule withholds data with high sample errors **or** high imputation
levels. A census removes the first branch, which is why census years sit lower.
The second branch remains, which is why they are not empty. All
`r sum(livestock$is_census_year & livestock$suppression_code == "s", na.rm = TRUE)`
census-year `S` cells in this table are Nelson.

So the 2012 spike is not a quality problem at all. All
`r sum(regions$year == 2012 & regions$suppressed)` withheld cells that year
carry `C`, not `S`. Separating the two flags turns an apparent anomaly into
the expected behaviour of a confidentiality rule that was replaced in 2017.

One qualification remains. I cannot tell from this extract whether the absence of
any flag before 2012 means no cell was withheld, or that the flag was not carried
in this export. The chart shows what the file contains.

One of the two regions this analysis has to drop is explained in the published
methodology rather than guessed at. Sampling error could not be calculated for
Nelson because only one responding unit was observed in each sampled stratum,
and at least two are needed to estimate a within-stratum variance. Auckland, the
other dropped region, is simply withheld, with no reason published.

### Independent recomputation

Stats NZ publishes a national sheep figure for June 2024. Recomputing it from the
regional cells in the same extract gives a different number, and the difference
is fully explained.

```{r}
#| label: tbl-recompute
r24 <- reconciliation |> filter(livestock_class == "Sheep", year == 2024)
data.frame(
  Quantity = c("Published, Total New Zealand, June 2024",
               "Recomputed as the sum of published regions",
               "Difference"),
  Head = c(fmt(r24$published), fmt(r24$region_sum), fmt(r24$residual))
) |> kable(align = "lr")
```

The difference is Auckland, whose 2024 value is suppressed. Filling it with zero
would have produced a total that looked tidier and was wrong by
`r fmt(r24$residual)` head.

## Method

1. **One table, one vintage.** Stats NZ table `AGR_AGR_003`, *Livestock Numbers
   by Regional Council*, dataflow `STATSNZ:AGR_AGR_003(1.0)`, final vintage,
   downloaded once and committed with its SHA-256. Full provenance is in
   [`data-raw/SOURCE.md`](https://github.com/Kenchch/nz-sheep-decline-by-region/blob/main/data-raw/SOURCE.md).
2. **Codes verified before code was written.** The extract carries numeric codes
   and no labels. Sheep, dairy cattle and beef cattle were each confirmed against
   the published June 2024 national figures before any analysis was run. The
   `LIVESTOCK` codelist holds 44 entries that nest inside one another; adding all
   44 for 2024 gives 115.7 million animals against a published 33.8 million. Only
   the three verified codes are used.
3. **Aggregates are separated, not summed.** `AREA` codes 10, 19 and 20 are the
   North Island, South Island and New Zealand totals. They are flagged in the
   analysis table and excluded from every regional sum. The extract publishes
   no area labels at all, only numeric codes; region names here follow the
   Statistical standard for geographic areas 2023 (SSGA23), including macrons,
   which differ from the older spellings the codelist still displays in
   Aotearoa Data Explorer.
4. **Suppressed cells stay suppressed.** A suppressed cell becomes `NA` and
   carries a flag. It is never filled with zero and never dropped. Regions
   missing either endpoint are excluded from the change calculation and counted
   in the text.
5. **Reconciliation is reported, not asserted.** No fixed tolerance is set
   between the regional sum and the published total. The residual is computed for
   every year and published above.
6. **Quantitative claims are computed, cited, or explicitly rounded — and the
   three are different things.** Numbers that come from the data are inline
   expressions evaluated at render time against the analysis table, never
   transcribed. Numbers that come from Stats NZ's published methodology —
   sample size, eligible population, response rate, imputation level, relative
   sampling error, the GST threshold, the codelist totals — *cannot* be
   computed from an aggregate extract; they are typed, and each is attributed
   under [Sources for the methodological
   statements](#sources-for-the-methodological-statements). Numbers in the
   plain-language layer at the top are deliberately rounded, with the exact
   value always available further down. Qualitative claims are stated with the
   scope over which they were checked. Earlier versions of this page each
   carried a universal claim that was typed rather than computed, and each was
   contradicted by a table printed directly beneath it; this is the rule that
   came out of that, including the version of this rule that claimed every
   number on the page was computed, which was itself false.

### Release schedule

This analysis uses the final vintage throughout. Up to and including the 2024
survey, Stats NZ published a provisional release in mid-December and a final
release in mid-May of the following year. From the 2025 survey those two were
replaced by a single final release in mid-April.

### Survey design

The reference period is the year ended 30 June. Censuses were held in 2002, 2007,
2012, 2017 and 2022, with annual sample surveys in the intervening years. The
sample is stratified by regional council area, ANZSIC06 group and size group. The
2025 survey drew 24,600 geographic locations from an estimated eligible
population of 45,800; note that a geographic location is the statistical unit and
is not the same thing as a farm.

The series starts at 2002 because Stats NZ states that the population for the
2002 census and subsequent surveys differs from that of earlier agricultural
censuses and surveys, and that figures from 2002 onward may not be directly
comparable with previous years. That is also why the 1994 observation carried in
the same table is excluded here. There was no agricultural production survey at
all in 1997, 1998 or 2001.

Values for non-respondents are imputed by random hot deck imputation, with the
imputation class formed from regional council area, ANZSIC06 group and production
data from previous years. This matters more than it may appear: for the 2025
survey Stats NZ reports that **30 percent** of the total sheep estimate was
imputed, against a relative sampling error of 3 percent at 95 percent confidence.
The response rate for the 2025 final release was 72.1 percent. The collection is
run by Stats NZ in partnership with the Ministry for Primary Industries.

#### The headline window is not design-matched

2002 is a census and 2025 is a sample survey, so the headline differences a
full-coverage count against an estimate that carries both sampling error and a
30 percent imputation rate. This is a common and accepted comparison — Stats NZ
publishes the series as one — but it is an asymmetry, and this report carries
the `is_census_year` flag through the analysis table precisely so that it can be
checked rather than assumed.

The census-to-census window that avoids the asymmetry is 2002 to 2022:

```{r}
#| label: tbl-matched
rbind(
  data.frame(Window = "2002–2025, census to sample survey (headline)",
             Regions = sensitivity$regions[sensitivity$start == 2002],
             Fall = fmt(sensitivity$fall[sensitivity$start == 2002]),
             `Top three share` = sprintf("%.1f%%", sensitivity$share[sensitivity$start == 2002]),
             `The three regions` = sensitivity$top3[sensitivity$start == 2002],
             check.names = FALSE),
  data.frame(Window = "2002–2022, census to census (design-matched)",
             Regions = matched$regions,
             Fall = fmt(matched$fall),
             `Top three share` = sprintf("%.1f%%", matched$share),
             `The three regions` = matched$top3,
             check.names = FALSE)
) |> kable(align = "lrrrl")
```

The design-matched window is measurable across one more region, because Auckland
is published in 2022 and withheld in 2025. It gives the same qualitative answer:
a fall of `r fmt(matched$fall/1e6, 1)` million head with the three largest
decliners accounting for `r sprintf("%.1f", matched$share)` percent.

The 2025 endpoint is used in the headline because it is the most recent
published figure, not because it is the better-matched one.

### What was excluded, and why

- **Territorial authority geography** — the surveys are designed to produce
  results to regional council level and the censuses to territorial authority
  level, so no annual series exists below the region.
- **Horticulture** — not collected in 2004, 2006, 2008, 2010, 2013, 2015, 2016,
  2018, 2021, 2023 or 2025, which is most of this window.
- **Forestry production** — Stats NZ no longer collects or compiles forestry
  production statistics; they are published by MPI. Stats NZ does still collect
  forestry *land use*. The 2025 release also notes that additional National
  Exotic Forest Description data has been received and that revisions to earlier
  land use and forestry statistics, especially for 2016 to 2021, are planned.
- **Pre-2002 data** — the population change described above, plus the absence of
  any survey in 1997, 1998 and 2001.
- **A choropleth map** — a sorted bar chart answers "which regions contributed
  most" and a map does not, because a map encodes land area rather than
  magnitude. Canterbury, Otago and Southland are also three of the largest
  regions by area, which is precisely the confound a map would introduce.
- **A second data set** — the question is answerable inside one table, and every
  extra source adds provenance to defend without adding evidence.

### Implications for output design

Three observations follow from the checks above, and they concern the shape of
the output rather than the sheep.

Regional council is the only geography available annually; territorial authority
exists only in census years, so any sub-regional time series is structurally
unavailable rather than merely unpublished. Suppression now runs above ten
percent of cells, and it is not spread evenly. Nelson and the Chatham Islands
account for `r sprintf("%.0f", 100 * sum(regions$suppressed & regions$region %in% c("Nelson", "Chatham Islands")) / sum(regions$suppressed))`
percent of every withheld cell in this table between them, Auckland for a
further `r sprintf("%.0f", 100 * sum(regions$suppressed & regions$region == "Auckland") / sum(regions$suppressed))`
percent, and
`r length(setdiff(unique(regions$region[regions$suppressed]), c("Nelson", "Chatham Islands", "Auckland")))`
other regions for the rest.

Two different grounds are at work, and it is worth separating them. Nelson and
the Chathams are withheld persistently — they hold the two smallest sheep
populations in the table — which makes their *trends* unusable. Auckland is
withheld only at scattered points including the 2025 endpoint, which is what
removes it from this analysis' *change* calculation while leaving most of its
series intact. Auckland is not a small region, and by sheep it is not even among
the smallest: its last published count is larger than Northland's or West
Coast's. Being withheld and being small are related here but not the same thing,
and an output that conflated them would mislead.

A suppression-aware grouping that pooled the persistently withheld regions would
preserve more signal than withholding them one at a time does. And the alternating census-and-survey design
makes every regional series in this table structurally uneven, which belongs on
the output rather than being left for a user to discover from a reconciliation
residual.

There is also a class of question this data cannot answer at all: anything at
farm level, or any breakdown finer than what is published. The access route for
those questions is approved microdata under the Five Safes framework, which I
have not used here.

### Sources for the methodological statements

Every statement about the collection above was read on the page cited here, not
carried over from secondary summaries.

| Statement | Source |
|---|---|
| Census years 2002, 2007, 2012, 2017, 2022; sample surveys in between; eligible population 45,800 and sample 24,600 geographic locations; stratification by regional council, ANZSIC06 group and size group; 2025 response rate 72.1 percent; sampling error and imputation levels by line code; GST registration level of \$60,000 giving "partial and unquantifiable" coverage below it; random hot deck imputation and imputation class; `C` before 2017 and input perturbation since; `S` for data with high sample errors or imputation levels; Nelson variance not estimable; horticulture years; forestry production now MPI; Tier 1 statistics | [Agricultural Production Statistics: June 2025 (Final), DataInfo+](https://datainfoplus.stats.govt.nz/item/nz.govt.stats/a9908d00-4827-46b2-a38a-310c0c75b029) |
| Programme run in partnership with MPI; provisional mid-December plus final mid-May up to 2024, single final mid-April from 2025; 2002 population differs from earlier collections; no survey in 1997, 1998, 2001; regional council annually and territorial authority in censuses | [Agriculture Production Statistics series, DataInfo+](https://datainfoplus.stats.govt.nz/item/nz.govt.stats/6362a469-f374-412e-ac25-d76fd0962003) |
| Release date of 16 April 2026 for the year to June 2025; NEFD data and planned revisions for 2016–2021 | [Agricultural production statistics: Year to June 2025](https://www.stats.govt.nz/information-releases/agricultural-production-statistics-year-to-june-2025/) |
| Mana Ōrite Relationship Agreement signed with the Data Iwi Leaders Group of the National Iwi Chairs Forum on 30 October 2019 | [Mana Ōrite Relationship Agreement](https://www.stats.govt.nz/about-us/what-we-do/mana-orite-relationship-agreement/) |

Statements I could not source on those pages are not made anywhere in this
report. Pages read 4 September 2026.

## Limitations

- Non-census years are **sample estimates**. I have not recomputed their sampling
  errors and I have applied no weighting of my own. A year-on-year movement in a
  single small region should not be read as real change.
- Suppressed cells are retained as `NA` and counted. Two regions are excluded
  from the 2002–2025 change entirely for that reason.
- The sheep-to-cattle comparisons are **ratios of head counts**. They are not
  claims about stocking rate, feed demand, land use or emissions.
- The frame is Stats NZ's Business Register, which is built on GST registration.
  The compulsory registration level is \$60,000, and Stats NZ describes coverage
  of units below that level as "partial and unquantifiable".
- **Imputation is large.** Stats NZ reports 30 percent of the 2025 total sheep
  estimate as imputed. Every regional figure in a survey year carries that, and
  this analysis does nothing to correct for it.
- Since 2017, confidentiality has been implemented by input perturbation, which
  adds noise at individual farm level. It is small relative to the effects
  discussed here, but it is present in every year from 2017 on.
- **The endpoints are of different collection designs.** 2002 is a census and
  2025 is a sample survey. The `is_census_year` flag is carried through the
  analysis table for exactly this reason, and the design-matched comparison is
  reported under [Survey design](#survey-design).
- **Choosing the start year is a choice, and the two halves of the headline
  survive it differently.** See the sensitivity analysis below.

### How much does the start year matter?

Recomputing the headline statistic from every start year available, against the
same 2025 endpoint:

```{r}
#| label: tbl-sensitivity
sensitivity |>
  transmute(
    `Start year` = ifelse(start %in% CENSUS_YEARS,
                          paste0(start, " (census)"), as.character(start)),
    `Regions measurable` = regions,
    `Fall (head)` = fmt(fall),
    `Top three share` = sprintf("%.1f%%", share),
    `The three regions` = top3
  ) |>
  kable(align = "lrrrl")
```

`Fall` here is measured across the regions listed as measurable in each window,
not against the published national total. The two differ by the withheld
regions: for 2002–2025 the regional figure is
`r fmt(sensitivity$fall[sensitivity$start == 2002])` head against a published
national fall of `r fmt(nat_drop)`, a gap of
`r sprintf("%.2f", 100 * (nat_drop - sensitivity$fall[sensitivity$start == 2002]) / sensitivity$fall[sensitivity$start == 2002])`
percent, as set out at the top of this page.

The concentration is robust. Across all
`r nrow(sensitivity)` start years the three largest-declining regions account
for between `r sprintf("%.0f", min(sensitivity$share))` and
`r sprintf("%.0f", max(sensitivity$share))` percent of the fall, and the range
is narrow enough that no start year would change the headline claim.

Which three regions they are is not robust. Manawatū-Whanganui is in the top
three in `r in_top3("Manawatū-Whanganui")` of the `r nrow(sensitivity)`
windows, Otago in `r in_top3("Otago")`, Southland in `r in_top3("Southland")`,
and Canterbury in only `r in_top3("Canterbury")` —
`r paste(sensitivity$start[vapply(strsplit(sensitivity$top3, ", ", fixed = TRUE), function(x) "Canterbury" %in% x, logical(1))], collapse = ", ")`.
Canterbury dominates the window this report uses because its decline is
concentrated in the early years; from a start year of 2009 or later it usually
drops out of the top three altogether.

That is worth stating plainly, because Canterbury is the region this report
names most often. It leads the 2002–2025 window, and it is the one region where
the substitution story carries weight. It is not a permanent feature of the
data.

## Ethics

This analysis uses published regional aggregates only. No microdata and no IDI
access is involved, so the Five Safes framework does not apply to it.

Stats NZ lists Tatauranga Umanga Māori among the internal users of this
collection, and signed the Mana Ōrite Relationship Agreement with the Data Iwi
Leaders Group of the National Iwi Chairs Forum in October 2019. Any Māori-specific
cut of this data therefore sits inside an established relationship between Stats
NZ and iwi Māori. It does not sit with an unaffiliated analyst working from
published aggregates, which is why this report does not attempt one and does not
characterise the terms of that agreement.

## Reproduce

Two steps, because they write different things.

```bash
Rscript R/checks.R    # regenerates outputs/*.csv: validation, coverage, reconciliation
quarto render         # regenerates this page and outputs/figures/*.png
```

`R/load.R` verifies the extract against the SHA-256 recorded in
`data-raw/SOURCE.md` before reading it, then tidies it; `R/checks.R` runs the
rules and both reconciliations; `index.qmd` produces this page and the figures.

Both sets of outputs are committed, so a clean checkout that runs the two
commands above finishes with no diff. If the extract were ever replaced without
`SOURCE.md` being updated, `R/load.R` stops rather than analysing it.

```{r}
#| label: session
#| code-fold: false
sessionInfo()
```
