---
title: "Where New Zealand's sheep went"
subtitle: "A regional breakdown of the flock decline, 2002–2025"
author: "Feng Jiang"
# Pinned rather than `today`: a render-time date would make every render
# produce a diff, which the reproducibility check would then report as a
# failure. Bump this, with the version in DESCRIPTION and CITATION.cff, when the
# analysis is materially revised.
date: 2026-09-23
lang: en-NZ
---

```{r setup}
#| include: false
# Every warning is an error. Most of the computation is in this hidden chunk
# and in inline expressions, whose warnings never reach the page, so a
# max() of an empty vector would otherwise render as "-Inf" and pass.
options(warn = 2)
source("R/checks.R")     # sources R/load.R, runs the rules and reconciles
source("R/analysis.R")   # pure helpers, tested in tests/test-r-analysis.R
suppressPackageStartupMessages({
  library(ggplot2); library(scales); library(knitr)
})

dir.create("outputs/figures", recursive = TRUE, showWarnings = FALSE)

VERSION <- unname(read.dcf("DESCRIPTION")[, "Version"])
CAPTION <- paste("Source: Stats NZ, Agricultural production statistics (AGR_AGR_003), CC BY 4.0.",
                 "Chart and calculations by Feng Jiang.", sep = "\n")

d       <- livestock
regions <- regions_only(d)
REGION_NAMES <- unname(AREA[setdiff(names(AREA), AREA_AGGREGATES)])
and_list <- function(x) {
  if (length(x) == 0) return("")
  if (length(x) == 1) return(as.character(x))
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))
}

# Sheep, first year to last year, by region -------------------------------
sheep_change <- endpoint_change(regions, "Sheep", START_YEAR, END_YEAR)
measured     <- sheep_change |> filter(!is.na(change))
headline     <- top3_for(regions, START_YEAR, END_YEAR)
measurable   <- headline$regions
excluded     <- sort(sheep_change$region[is.na(sheep_change$change)])
decline_sum  <- -headline$fall
top3         <- head(measured |> filter(change < 0), 3)
fourth       <- (measured |> filter(change < 0))[4, ]
stopifnot("the prose compares the third and fourth largest declines" =
            nrow(fourth) == 1L, !is.na(fourth$change))
top3_share   <- headline$share
# The same share with the fourth-largest decline counted third instead.
alt_share    <- 100 * (sum(top3$change[1:2]) + fourth$change) / decline_sum
top3_base    <- headline$base_share

# Each region's share of the fall against its share of the starting flock. If
# every region had fallen by the same proportion the two would be equal, so
# the gap is the part of the headline that is about the decline rather than
# about which regions were largest to begin with.
measured <- measured |>
  mutate(base_share = 100 * start / sum(start),
         fall_share = 100 * change / decline_sum,
         excess_pp  = fall_share - base_share)
measured_rate <- 100 * -sum(measured$change) / sum(measured$start)
over_share    <- top3$region[top3$region %in%
                               measured$region[measured$excess_pp > 0]]
under_share   <- setdiff(top3$region, over_share)
steepest      <- measured |> arrange(pct) |> head(4)

stopifnot("the headline assumes every measurable region declined" =
            all(measured$change < 0))

nat_sheep <- national(d) |> filter(livestock_class == "Sheep")
nat_start <- nat_sheep$head[nat_sheep$year == START_YEAR]
nat_end   <- nat_sheep$head[nat_sheep$year == END_YEAR]
nat_drop  <- nat_start - nat_end
# The last year the flock was still at or above its starting level, so the
# figure 4 text says where the fall begins rather than "throughout".
sheep_flat_until <- max(nat_sheep$year[nat_sheep$head >= nat_start])
nat_pct   <- 100 * nat_drop / nat_start
top3_national_share <- 100 * -sum(top3$change) / nat_drop
national_gap_pct <- 100 * (nat_drop + decline_sum) / -decline_sum

# Stats NZ's published relative sampling error for the national sheep
# estimate, 2025 survey, 95 percent confidence (cited under Sources). The
# 2002 endpoint is a census, so this is the uncertainty of the end alone.
RSE_2025 <- 0.03
half_width <- RSE_2025 * nat_end

nat_change <- function(class, y0, y1) {
  x <- national(d) |> filter(livestock_class == class)
  x$head[x$year == y1] - x$head[x$year == y0]
}
dairy_gain <- nat_change("Dairy cattle", START_YEAR, END_YEAR)
beef_gain  <- nat_change("Beef cattle", START_YEAR, END_YEAR)

# Substitution, all three classes ----------------------------------------
substitution <- bind_rows(lapply(unname(LIVESTOCK), function(class)
  endpoint_change(regions, class, START_YEAR, END_YEAR) |>
    transmute(region, livestock_class = class, change))) |>
  pivot_wider(names_from = livestock_class, values_from = change)

top5_regions <- head(measured$region, 5)
per_100 <- function(region, class) {
  x <- substitution[match(region, substitution$region), ]
  -100 * x[[class]] / x$Sheep
}
# Dairy cattle gained per 100 sheep lost, largest first, among the five
# largest sheep declines. The prose names the first two and the largest of
# the rest, so the gap between them is visible rather than asserted.
dairy_ratio <- sort(setNames(per_100(top5_regions, "Dairy cattle"), top5_regions),
                    decreasing = TRUE)

# The endpoint count hides the shape of the series: dairy and sheep moved in
# opposite directions to the 2014 dairy peak and in the same direction after
# it. Counting per window is what makes that visible.
win_full  <- dairy_window_counts(regions, START_YEAR, END_YEAR)
win_early <- dairy_window_counts(regions, START_YEAR, DAIRY_PEAK_YEAR)
win_late  <- dairy_window_counts(regions, DAIRY_PEAK_YEAR, END_YEAR)
# The three windows admit different regions, so the counts are also given on
# the regions measurable in all three.
common <- Reduce(intersect, list(win_full$regions, win_early$regions,
                                 win_late$regions))
fell_in <- function(y0, y1, keep) {
  ch <- endpoint_change(regions, "Dairy cattle", y0, y1)
  sum(ch$change[ch$region %in% keep] < 0)
}
window_counts <- data.frame(
  y0 = c(START_YEAR, START_YEAR, DAIRY_PEAK_YEAR),
  y1 = c(END_YEAR, DAIRY_PEAK_YEAR, END_YEAR)
)
window_counts$n    <- c(win_full$n, win_early$n, win_late$n)
window_counts$fell <- c(win_full$fell, win_early$fell, win_late$fell)
window_counts$excluded <- vapply(list(win_full, win_early, win_late), function(w)
  and_list(setdiff(REGION_NAMES, w$regions)), character(1))
window_counts$common_fell <- mapply(fell_in, window_counts$y0, window_counts$y1,
                                    MoreArgs = list(keep = common))
chathams_zero_late <- win_late$zero_both == 1 &&
  "Chatham Islands" %in% win_late$regions

# National dairy against its peak, for the same reason.
nat_dairy  <- national(d) |> filter(livestock_class == "Dairy cattle")
dairy_start <- nat_dairy$head[nat_dairy$year == START_YEAR]
dairy_peak  <- nat_dairy$head[nat_dairy$year == DAIRY_PEAK_YEAR]
dairy_rise  <- dairy_peak - dairy_start
dairy_fall  <- dairy_peak - nat_dairy$head[nat_dairy$year == END_YEAR]
stopifnot("DAIRY_PEAK_YEAR is described as the in-sample dairy peak" =
            nat_dairy$year[which.max(nat_dairy$head)] == DAIRY_PEAK_YEAR)
dairy_runner_up <- nat_dairy |> filter(year != DAIRY_PEAK_YEAR) |>
  slice_max(head, n = 1)

sheep_peak_year <- nat_sheep$head[nat_sheep$year == DAIRY_PEAK_YEAR]
sheep_early_drop <- nat_start - sheep_peak_year
sheep_late_drop  <- sheep_peak_year - nat_end
annual_rate <- function(a, b, years) 100 * (1 - (b / a)^(1 / years))
sheep_early_rate <- annual_rate(nat_start, sheep_peak_year, DAIRY_PEAK_YEAR - START_YEAR)
sheep_late_rate  <- annual_rate(sheep_peak_year, nat_end, END_YEAR - DAIRY_PEAK_YEAR)

# Regions among the top five that lost at least one cattle class as well as
# sheep, named from the data rather than typed into the prose.
lost_cattle_class <- substitution |>
  filter(region %in% top5_regions, `Dairy cattle` < 0 | `Beef cattle` < 0) |>
  pull(region) |> sort()

# Endpoint sensitivity ----------------------------------------------------
# The same headline statistic recomputed from every start year, and to every
# end year, that leaves a window of at least MIN_WINDOW years. The five-year
# minimum is an analysis choice to focus on longer windows.
MIN_WINDOW <- 5L
LAST_START <- END_YEAR - MIN_WINDOW
FIRST_END  <- START_YEAR + MIN_WINDOW

excluded_regions <- function(y0, y1) {
  ch <- endpoint_change(regions, "Sheep", y0, y1)
  setdiff(REGION_NAMES, ch$region[!is.na(ch$change)])
}
excluded_in <- function(y0, y1) and_list(excluded_regions(y0, y1))
sensitivity <- do.call(rbind, lapply(START_YEAR:LAST_START, function(y0)
  top3_for(regions, y0, END_YEAR)))
end_sensitivity <- do.call(rbind, lapply(FIRST_END:END_YEAR, function(y1)
  top3_for(regions, START_YEAR, y1)))
end_sensitivity$excluded <- vapply(end_sensitivity$end, function(y1)
  excluded_in(START_YEAR, y1), character(1))
stopifnot("the prose says the share stays above half for every end year" =
            all(end_sensitivity$share > 50))

# Design-matched comparison: census start, census end.
matched <- top3_for(regions, START_YEAR, LAST_CENSUS)
matched_top3 <- strsplit(matched$top3, ", ", fixed = TRUE)[[1]]
dropped_in_matched <- setdiff(top3$region, matched_top3)
added_in_matched   <- setdiff(matched_top3, top3$region)
matched_gain <- setdiff(excluded, excluded_regions(START_YEAR, LAST_CENSUS))
stopifnot("the matched-window prose assumes the top three differ" =
            length(added_in_matched) == 1L)

# Withheld and absent cells -----------------------------------------------
# Every expected region-class-year cell, published, flagged or absent. Before
# 2012 the export omits cells rather than flagging them.
grid <- cell_status_grid(regions, EXPECTED_YEARS, unname(LIVESTOCK), REGION_NAMES)
cells_per_year <- length(REGION_NAMES) * length(LIVESTOCK)
status_by_year <- grid |>
  count(year, status) |>
  mutate(rate = 100 * n / cells_per_year)
withheld_rate <- grid |>
  group_by(year) |>
  summarise(rate = 100 * sum(status %in% c("C", "S")) / n(), .groups = "drop")

c_years <- sort(unique(grid$year[grid$status == "C"]))
s_years <- sort(unique(grid$year[grid$status == "S"]))
absent_years <- sort(unique(grid$year[grid$status == "Absent"]))
first_flag_year <- min(c(c_years, s_years))
stopifnot("the absent-row narrative assumes rows stop being absent when flags start" =
            max(absent_years) < first_flag_year)
peak_row <- withheld_rate |> slice_max(rate, n = 1)
recent <- recent_above(withheld_rate$rate, 10, 4)
stopifnot("the prose says S appears every year from 2017" =
            all(2017:END_YEAR %in% s_years))
n_pass <- sum(results$fails == 0)
number_word <- function(n) c("None", "One", "Two", "Three", "Four", "Five")[n + 1]

# Class-years whose region row is incomplete although nothing is flagged: the
# only way a cell could be missing before 2012.
unflagged_gaps <- residual_tiers |>
  filter(regions_present < N_REGIONS, regions_suppressed == 0)
gap_nonzero <- sum(unflagged_gaps$residual != 0)

# The 2012 confidentiality cells across every livestock code in the extract,
# and how many of the ten earlier years each one was absent from the export.
c_2012 <- raw |>
  filter(year_agr_agr_003 == "2012", obs_status == "c",
         !area_agr_agr_003 %in% AREA_AGGREGATES) |>
  distinct(livestock_agr_agr_003, area_agr_agr_003)
earlier <- raw |>
  filter(year_agr_agr_003 %in% as.character(START_YEAR:2011)) |>
  count(livestock_agr_agr_003, area_agr_agr_003, name = "present")
c_2012 <- c_2012 |>
  left_join(earlier, by = c("livestock_agr_agr_003", "area_agr_agr_003")) |>
  mutate(present = coalesce(present, 0L),
         absent  = length(START_YEAR:2011) - present)

s_census <- grid |> filter(status == "S", year %in% CENSUS_YEARS)
s_by_year <- grid |> filter(status == "S") |> count(year)
s_count <- function(y) sum(s_by_year$n[s_by_year$year == y])
s_census_regions <- sort(unique(s_census$region))

# West Coast against the Chathams, by the metric the sentence names.
sheep_mean <- regions |>
  filter(livestock_class == "Sheep", !is.na(head)) |>
  group_by(region) |>
  summarise(mean_head = mean(head), years = n(), .groups = "drop") |>
  arrange(mean_head)

# Codelist nesting, computed from the extract rather than typed.
codes_2024 <- raw |>
  filter(area_agr_agr_003 == "20", year_agr_agr_003 == "2024") |>
  mutate(head = as.numeric(obs_value))
all_codes_2024   <- sum(codes_2024$head, na.rm = TRUE)
three_codes_2024 <- sum(codes_2024$head[codes_2024$livestock_agr_agr_003 %in%
                                          names(LIVESTOCK)])

# Residual tiers -----------------------------------------------------------
off_tier <- residual_tiers |> filter(tier == "Fully published, off by a few head")
stopifnot("the tier prose needs every tier to be populated" =
            nrow(off_tier) > 0, any(residual_tiers$tier == "Exact"),
          any(residual_tiers$tier == "Incomplete region row" & residual_tiers$residual != 0))
fp_recent <- residual_tiers |> filter(fully_published, year >= 2017)
next_smallest <- min(abs(residual_tiers$residual[
  residual_tiers$tier == "Incomplete region row" & residual_tiers$residual != 0]))

# Key figures, for the README ---------------------------------------------
# The README repeats a handful of these numbers. Each is written here with the
# exact phrase the README must contain, and tests/test-readme-figures.R fails
# if the README stops containing any of them, so the landing page cannot drift
# from the report the way it once did.
kf <- function(name, value, phrase) data.frame(name = name, value = value, readme = phrase)
key_figures <- rbind(
  kf("national_fall", fmt(nat_drop / 1e6, 1), sprintf("about %s million head", fmt(nat_drop / 1e6, 1))),
  kf("national_fall_pct", fmt(nat_pct, 0), sprintf("(%s percent)", fmt(nat_pct, 0))),
  kf("sampling_half_width", fmt(half_width / 1e6, 1), sprintf("±%s million", fmt(half_width / 1e6, 1))),
  kf("measurable_regions", measurable, sprintf("%d regions measurable at both ends", measurable)),
  kf("top3", headline$top3, and_list(top3$region)),
  kf("top3_share", fmt(top3_share, 0), sprintf("**%s percent** of the fall", fmt(top3_share, 0))),
  kf("top3_base_share", fmt(top3_base, 1), sprintf("%s percent of the %d flock", fmt(top3_base, 1), START_YEAR)),
  kf("fourth", fourth$region, sprintf("%s and %s", top3$region[3], fourth$region)),
  kf("third_gap", fmt(headline$third_gap), sprintf("%s head apart", fmt(headline$third_gap))),
  kf("dairy_rise", fmt(dairy_rise / 1e6, 2),
     sprintf("+%s million (%s percent)", fmt(dairy_rise / 1e6, 2), fmt(100 * dairy_rise / dairy_start, 1))),
  kf("dairy_fall", fmt(dairy_fall / 1e6, 2), sprintf("−%s million", fmt(dairy_fall / 1e6, 2))),
  kf("dairy_net", fmt(dairy_gain / 1e3, 0), sprintf("%s thousand net", fmt(dairy_gain / 1e3, 0))),
  kf("dairy_count_early", paste(win_early$fell, "of", win_early$n),
     sprintf("**%d of %d** regions over %d–%d", win_early$fell, win_early$n, START_YEAR, DAIRY_PEAK_YEAR)),
  kf("dairy_count_late", paste(win_late$fell, "of", win_late$n),
     sprintf("**%d of %d** over %d–%d", win_late$fell, win_late$n, DAIRY_PEAK_YEAR, END_YEAR)),
  kf("dairy_count_full", paste(win_full$fell, "of", win_full$n),
     sprintf("%d of %d over the whole period", win_full$fell, win_full$n)),
  kf("matched_fall", fmt(matched$fall / 1e6, 1), sprintf("%s million head", fmt(matched$fall / 1e6, 1))),
  kf("matched_share", fmt(matched$share, 1), sprintf("top three at %s percent", fmt(matched$share, 1))),
  kf("top3_national_share", fmt(top3_national_share, 1),
     sprintf("%s percent against the published national fall", fmt(top3_national_share, 1))),
  kf("national_gap_pct", fmt(national_gap_pct, 2), sprintf("%s percent larger", fmt(national_gap_pct, 2))),
  kf("excluded", and_list(excluded), sprintf("%s are withheld in %d", and_list(excluded), END_YEAR)),
  kf("census_rule_fails", results$fails[results$rule == "census_year_reported"],
     sprintf("fires %d times", results$fails[results$rule == "census_year_reported"])),
  kf("tiers", paste(table(residual_tiers$tier)[c("Exact", "Fully published, off by a few head",
                                                  "Incomplete region row")], collapse = "/"),
     sprintf("%d exact, %d off by at most %d head, %d incomplete",
             sum(residual_tiers$tier == "Exact"), nrow(off_tier),
             max(abs(off_tier$residual)),
             sum(residual_tiers$tier == "Incomplete region row"))),
  kf("unflagged_gap_nonzero", paste(gap_nonzero, "of", nrow(unflagged_gaps)),
     sprintf("%d of the %d", gap_nonzero, nrow(unflagged_gaps))),
  kf("sensitivity_range",
     paste0(fmt(min(sensitivity$share), 0), "–", fmt(max(sensitivity$share), 0)),
     sprintf("%s–%s percent", fmt(min(sensitivity$share), 0), fmt(max(sensitivity$share), 0)))
)
readr::write_csv(key_figures, "outputs/key-figures.csv")
```

::: {.callout-note appearance="simple"}
This work is based on Stats NZ's data (*Agricultural production statistics*,
table `AGR_AGR_003`), licensed by Stats NZ for re-use under the [Creative
Commons Attribution 4.0 International
licence](https://creativecommons.org/licenses/by/4.0/). The tables and figures
here are aggregated and reshaped from that data. Version `r VERSION`; see the
[changelog](https://github.com/Kenchch/nz-sheep-decline-by-region/blob/main/CHANGELOG.md)
for what changed between versions.
:::

## The short version

New Zealand's sheep flock fell by **about `r fmt(nat_drop/1e6, 1)` million** head
between June `r START_YEAR` and June `r END_YEAR`, a drop of `r fmt(nat_pct, 0)`
percent. The `r END_YEAR` figure is a sample-survey estimate: at Stats NZ's
published relative sampling error of `r RSE_2025 * 100` percent, the end point alone
is uncertain by about ±`r fmt(half_width/1e6, 1)` million head.

Three points.

- The fall is spread across the country roughly in proportion to where the sheep
  were. `r and_list(top3$region)` held `r fmt(top3_base, 1)` percent of the
  `r START_YEAR` flock in the `r measurable` regions where both endpoints are
  published, and account for `r fmt(top3_share, 0)` percent of the fall there —
  `r fmt(top3_share - top3_base, 1)` percentage points more than their size
  alone would give. Of the three, only `r and_list(over_share)` fell by more than
  their share of the flock.
  `r top3$region[3]` is third by the narrowest of margins: it lost
  `r fmt(-top3$change[3])` head against `r fourth$region`'s
  `r fmt(-fourth$change)`, a difference of `r fmt(headline$third_gap)` head.
- Over the same `r END_YEAR - START_YEAR` years the national dairy herd rose by
  **`r fmt(dairy_gain/1e3, 0)` thousand** head net, while beef cattle *fell* by
  `r fmt(-beef_gain/1e3, 0)` thousand. The dairy net hides a
  `r fmt(dairy_rise/1e6, 2)` million rise to `r DAIRY_PEAK_YEAR` and a
  `r fmt(dairy_fall/1e6, 2)` million fall since.
- Whether dairy fell *alongside* sheep depends on the window: in
  `r win_full$fell` of the `r win_full$n` regions with sheep and dairy published at
  both ends for `r START_YEAR`–`r END_YEAR`, in only `r win_early$fell` of
  `r win_early$n` for `r START_YEAR`–`r DAIRY_PEAK_YEAR`, and in
  `r win_late$fell` of `r win_late$n` for `r DAIRY_PEAK_YEAR`–`r END_YEAR`.
  The regional counts for all three windows are in
  [Three endpoint windows](#three-endpoint-windows).

The regional shares use the measurable regional fall as their denominator.
Against the published national fall, the same three regions account for
`r fmt(top3_national_share, 1)` percent. The national fall is
`r fmt(national_gap_pct, 2)` percent larger than the measurable regional fall
because `r and_list(excluded)` have withheld `r END_YEAR` values. The ±`r fmt(half_width/1e6, 1)` million sampling
margin above is national and belongs to one endpoint; it says nothing about the gap between two regions.
Stats NZ does not publish regional sampling errors in this table, so no
uncertainty is attached to the regional shares or ranks. `r top3$region[3]` and
`r fourth$region` differ by `r fmt(100 * headline$third_gap / top3$end[3], 1)`
percent of `r top3$region[3]`'s `r END_YEAR` flock and are not ranked against each
other here. The top-three share depends only on which of the two is counted
third: `r fmt(top3_share, 1)` percent with `r top3$region[3]`,
`r fmt(alt_share, 1)` percent with `r fourth$region`.

::: {.callout-tip collapse="true"}
## Terms used on this page

- **Head** — one animal. Every count here is a head count, not a stock unit.
- **Class** — one of the three livestock totals used: sheep (code 6731), dairy
  cattle (7193) and beef cattle (7077).
- **Class-year** — one class in one year: `r length(LIVESTOCK)` classes ×
  `r length(EXPECTED_YEARS)` years = `r length(LIVESTOCK) * length(EXPECTED_YEARS)`.
- **Withheld** — a cell present in the export with no value and a flag: `C`
  (confidentiality) or `S` (quality suppression). The data column for this is
  called `suppressed`.
- **Absent** — a region-class-year with no row in the export at all.
- **Endpoint** — the first or last year of a comparison window.
- **Measurable region** — a region published (neither withheld nor absent) at
  both endpoints of the window in question.
- **Residual** — the published national total minus the sum of the published
  regions, for one class-year.
- **Tier** — a class-year's residual classified by whether its region row is
  complete; see [Data quality checks](#data-quality-checks).
:::

### What this means

The common story is that sheep farms became dairy farms. In head counts, in this
table, over the `r START_YEAR`–`r END_YEAR` endpoints, dairy cattle rose by
**`r fmt(dairy_gain/1e3, 0)` thousand** while sheep fell by
**`r fmt(nat_drop/1e6, 1)` million**.

That net figure hides the shape of the series. The national dairy herd rose by
`r fmt(dairy_rise/1e6, 2)` million (`r fmt(100*dairy_rise/dairy_start, 1)` percent)
to its `r DAIRY_PEAK_YEAR` peak and has fallen by `r fmt(dairy_fall/1e6, 2)` million
since. Over `r START_YEAR`–`r DAIRY_PEAK_YEAR`, the period with the larger
absolute sheep decline (−`r fmt(sheep_early_drop/1e6, 1)` million, against
−`r fmt(sheep_late_drop/1e6, 1)` million since; the annual rates are similar, at
`r fmt(sheep_early_rate, 2)` and `r fmt(sheep_late_rate, 2)` percent a year), the
two moved in opposite directions. Over `r DAIRY_PEAK_YEAR`–`r END_YEAR` both fell.
`r DAIRY_PEAK_YEAR` is where the dairy herd peaks in this sample, and it was
chosen after looking at the series: splitting at a peak maximises the
before/after contrast by construction. `r dairy_runner_up$year` is within
`r fmt(100 * (dairy_peak - dairy_runner_up$head) / dairy_peak, 1)` percent of
it.

Whether the earlier dairy growth *replaced* the sheep in feed terms cannot be
settled from head counts — see the stock-unit note below.

### What these numbers do not tell you

They are head counts. They say nothing about land area, feed demand, grazing
pressure, farm profitability, or what the land is used for now. A dairy cow and a
sheep are not exchangeable units, and this page does not treat them as such.

## The regional breakdown

```{r}
#| label: fig-contribution
#| fig-cap: !expr 'sprintf("Sheep declines across the %d regions with both endpoints published, June %d to June %d. %s are not shown: their %d values are withheld.", measurable, START_YEAR, END_YEAR, and_list(excluded), END_YEAR)'
#| fig-alt: !expr 'sprintf("Horizontal bar chart of %d regions ordered by how many sheep they lost between %d and %d. %s is largest at %s million head fewer, then %s at %s million and %s at %s million; these three, drawn darker, make up %s percent of the fall across these regions and held %s percent of their %d flock. %s is fourth at %s million. The remaining regions range down to %s at %s million. A tick on each bar marks the fall the region would have had at the %s percent average rate.", measurable, START_YEAR, END_YEAR, top3$region[1], fmt(-top3$change[1]/1e6, 2), top3$region[2], fmt(-top3$change[2]/1e6, 2), top3$region[3], fmt(-top3$change[3]/1e6, 2), fmt(top3_share, 0), fmt(top3_base, 1), START_YEAR, fourth$region, fmt(-fourth$change/1e6, 2), tail(measured$region, 1), fmt(-tail(measured$change, 1)/1e6, 2), fmt(measured_rate, 1))'
plot_dat <- measured |>
  mutate(region = factor(region, levels = rev(region)),
         millions = -change / 1e6,
         proportional = measured_rate / 100 * start / 1e6,
         group = ifelse(region %in% top3$region, "Top three", "Other regions"))

p1 <- ggplot(plot_dat, aes(x = millions, y = region)) +
  geom_col(aes(fill = group), width = 0.72) +
  geom_errorbar(aes(xmin = proportional, xmax = proportional),
                width = 0.72, colour = "grey15", linewidth = 0.6) +
  geom_text(aes(x = pmax(millions, proportional), label = sprintf("%.2f", millions)),
            hjust = -0.15, size = 3.1, colour = "grey25") +
  annotate("label", x = max(plot_dat$millions) * 0.9,
           y = nrow(plot_dat) - 1.9,
           label = sprintf("%s:\n%.0f%% of the fall across these %d regions,\n%.1f%% of their %d flock",
                           and_list(top3$region), top3_share, measurable,
                           top3_base, START_YEAR),
           hjust = 0, size = 3.2, linewidth = 0, fill = "grey96", colour = "grey20") +
  scale_fill_manual(values = c("Top three" = "#1b4332", "Other regions" = "#95a5a0"),
                    guide = "none") +
  scale_x_continuous(labels = label_comma(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.3))) +
  labs(
    title    = sprintf("Three regions account for %.0f%% of the measurable regional fall,\nand held %.1f%% of the %d flock in these regions",
                       top3_share, top3_base, START_YEAR),
    subtitle = sprintf("Reduction in sheep, millions of head, June %d to June %d. Tick: fall at the %.1f%% average rate",
                       START_YEAR, END_YEAR, measured_rate),
    x = "Million head fewer", y = NULL,
    caption  = paste0(and_list(excluded), " not shown: ", END_YEAR,
                      " value withheld.\n", CAPTION)
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title.position = "plot",
        plot.title    = element_text(face = "bold", size = 12),
        plot.caption  = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/01-regional-contribution.png", p1,
       width = 9, height = 6, dpi = 200, bg = "white")
p1
```

`r and_list(excluded)` are missing from that chart: each has a withheld
`r END_YEAR` value, so no change can be computed for them. They are excluded
and counted, never imputed and never set to zero.

By percentage rather than head, the steepest declines are elsewhere:
`r and_list(sprintf("%s (%s percent)", steepest$region, fmt(-steepest$pct, 1)))`,
against `r fmt(measured_rate, 1)` percent across all measurable regions.

The share of the fall taken by the three largest declines stays above half for
every end year tested below; which three they are does not stay the same. `r top3$region[3]` and `r fourth$region` are
all but tied, and the census-to-census window below puts
`r and_list(added_in_matched)` in the top three in place of
`r and_list(dropped_in_matched)`.

```{r}
#| label: tbl-share-vs-size
#| tbl-cap: !expr 'sprintf("Each measurable region''s share of the %d–%d sheep fall against its share of the %d flock.", START_YEAR, END_YEAR, START_YEAR)'
measured |>
  transmute(Region = region,
            `Share of fall` = sprintf("%.1f%%", fall_share),
            `Share of flock` = sprintf("%.1f%%", base_share),
            `Difference (points)` = sprintf("%+.1f", excess_pp),
            `Own decline` = sprintf("%.1f%%", -pct)) |>
  kable(align = "lrrrr")
```

## Did dairy replace the sheep?

In head counts, among the five regions that lost the most sheep, only
`r names(dairy_ratio)[1]` and, to a lesser extent, `r names(dairy_ratio)[2]`
added dairy cattle in numbers that register against their sheep losses:
`r fmt(dairy_ratio[1], 1)` and `r fmt(dairy_ratio[2], 1)` dairy cattle per 100
sheep lost, against at most `r fmt(dairy_ratio[3], 1)` in the other three.
Nationally, dairy grew until `r DAIRY_PEAK_YEAR` and has fallen since. Whether
dairy replaced sheep in feed or land terms cannot be answered from this table.

```{r}
#| label: fig-national
#| fig-cap: !expr 'sprintf("National sheep, dairy cattle and beef cattle, indexed to %d = 100, June %d to June %d.", START_YEAR, START_YEAR, END_YEAR)'
#| fig-alt: !expr 'sprintf("Three line panels, one per class, each indexed to %d equals 100. Sheep stays at or above its %d level until %d, then falls to %s by %d. Dairy cattle rises to %s at its %d peak, marked by a dashed line, then falls to %s by %d. Beef cattle ends at %s. Census years are marked with filled points.", START_YEAR, START_YEAR, sheep_flat_until, fmt(100 * nat_end / nat_start, 0), END_YEAR, fmt(100 * dairy_peak / dairy_start, 0), DAIRY_PEAK_YEAR, fmt(100 * nat_dairy$head[nat_dairy$year == END_YEAR] / dairy_start, 0), END_YEAR, fmt(100 * (1 + beef_gain / national(d)$head[national(d)$livestock_class == "Beef cattle" & national(d)$year == START_YEAR]), 0))'
index_dat <- national(d) |>
  group_by(livestock_class) |>
  mutate(index = 100 * head / head[year == START_YEAR]) |>
  ungroup() |>
  mutate(livestock_class = factor(livestock_class,
                                  levels = c("Sheep", "Dairy cattle", "Beef cattle")))

p4 <- ggplot(index_dat, aes(x = year, y = index)) +
  geom_hline(yintercept = 100, colour = "grey70") +
  geom_vline(xintercept = DAIRY_PEAK_YEAR, colour = "grey50", linetype = "dashed") +
  geom_line(colour = "#2d6a4f", linewidth = 0.9) +
  geom_point(data = function(x) filter(x, is_census_year),
             colour = "#2d6a4f", size = 2) +
  facet_wrap(~ livestock_class, nrow = 1) +
  scale_x_continuous(breaks = c(START_YEAR, CENSUS_YEARS[-1], END_YEAR)) +
  labs(
    title    = sprintf("Sheep fell %s%% overall; dairy rose to its %d peak and has fallen since",
                       fmt(nat_pct, 0), DAIRY_PEAK_YEAR),
    subtitle = sprintf("National head count, index %d = 100. Points: census years. Dashed line: %d dairy peak",
                       START_YEAR, DAIRY_PEAK_YEAR),
    x = NULL, y = sprintf("Index, %d = 100", START_YEAR),
    caption  = CAPTION
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/04-national-index.png", p4,
       width = 9, height = 4.5, dpi = 200, bg = "white")
p4
```

### Three endpoint windows

```{r}
#| label: tbl-window-counts
#| tbl-cap: "Regions in which dairy cattle fell, by window, among regions with sheep and dairy published at both ends."
window_counts |>
  transmute(Window = paste(y0, y1, sep = "–"),
            `Dairy fell` = paste(fell, "of", n),
            `Not measurable` = excluded,
            !!sprintf("Dairy fell, the %d regions measurable in all three", length(common)) :=
              paste(common_fell, "of", length(common))) |>
  kable(align = "lrlr")
```

Each window admits the regions published at both of its own endpoints, so the
three counts are not over the same regions; the last column repeats them on the
`r length(common)` regions measurable in all three.
`r if (chathams_zero_late) sprintf("The %d–%d count includes the Chatham Islands, which had no dairy cattle at either end and so could not have fallen.", DAIRY_PEAK_YEAR, END_YEAR) else ""`

```{r}
#| label: tbl-dairy-windows
#| tbl-cap: "Published national totals for three comparison windows."
class_windows |>
  transmute(Window = paste(start_year, end_year, sep = "–"),
            Design = paste(start_design, end_design, sep = " → "),
            Class = livestock_class,
            `Start head` = fmt(start_head), `End head` = fmt(end_head),
            `Change (head)` = sprintf("%s%s", ifelse(change_head > 0, "+", ""),
                                      fmt(change_head)),
            `Change (%)` = sprintf("%+.1f", change_pct)) |>
  kable(align = "lllrrrr")
```

These are published **national totals**, not sums with missing regions filled
with zero. Splitting at `r DAIRY_PEAK_YEAR` separates the earlier dairy expansion
from the later period; the `r START_YEAR`–`r LAST_CENSUS` window compares two
censuses. The other windows include sample-survey estimates, so their
differences also carry sampling error.

A head of dairy cattle and a sheep do not require the same feed, so a head-count
comparison is not a feed-equivalent one. This analysis uses total-head codes
only: total sheep (6731), total dairy cattle (7193) and total beef cattle
(7077). The comparisons in this section are in head counts. Whether they also
hold in feed terms — stock units — is not asserted here either way: a
class-specific stock-unit comparison was not attempted.

Doing one properly needs two things this analysis does not have. It needs the
class-specific codes rather than the totals, because applying one adult-cow or
adult-ewe coefficient to a mixed-age total assumes a herd composition the
extract does not state. And it needs stock-unit coefficients from a citable
source, matched to those classes. The extract carries numeric codes and no
labels, so identifying the class-specific codes is itself a verification step,
not a lookup.

```{r}
#| label: fig-substitution
#| fig-cap: !expr 'sprintf("Change in sheep, dairy cattle and beef cattle in the five regions that lost the most sheep, June %d to June %d. Head counts are not feed-equivalent.", START_YEAR, END_YEAR)'
#| fig-alt: !expr 'sprintf("Dot chart for the five regions with the largest sheep declines, in millions of head. In each region the sheep marker sits between %s and %s million head below zero. The dairy and beef markers are much closer to zero: the largest dairy gains are %s at %s million and %s at %s million. %s lost at least one cattle class as well as sheep.", fmt(min(-substitution$Sheep[substitution$region %in% top5_regions])/1e6, 2), fmt(max(-substitution$Sheep[substitution$region %in% top5_regions])/1e6, 2), names(dairy_ratio)[1], fmt(substitution$`Dairy cattle`[substitution$region == names(dairy_ratio)[1]]/1e6, 2), names(dairy_ratio)[2], fmt(substitution$`Dairy cattle`[substitution$region == names(dairy_ratio)[2]]/1e6, 2), and_list(lost_cattle_class))'
classes <- c("Sheep", "Dairy cattle", "Beef cattle")
offset  <- c("Sheep" = 0.22, "Dairy cattle" = 0, "Beef cattle" = -0.22)
dumb <- substitution |>
  filter(region %in% top5_regions) |>
  pivot_longer(-region, names_to = "livestock_class", values_to = "change") |>
  mutate(
    y = match(region, rev(top5_regions)) + offset[livestock_class],
    livestock_class = factor(livestock_class, levels = classes),
    millions = change / 1e6
  )

p2 <- ggplot(dumb, aes(x = millions, y = y)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_segment(aes(x = 0, xend = millions, yend = y, colour = livestock_class),
               linewidth = 0.6, alpha = 0.6) +
  # Shape as well as colour, so colour is not the only channel carrying the
  # distinction. The green and the red-brown are close under red-green colour
  # vision deficiency; WCAG 1.4.1 asks that colour not be used alone.
  geom_point(aes(colour = livestock_class, shape = livestock_class), size = 3.2) +
  scale_colour_manual(values = c("Sheep" = "#2d6a4f",
                                 "Dairy cattle" = "#c1666b",
                                 "Beef cattle" = "#4a6fa5"), name = NULL) +
  scale_shape_manual(values = c("Sheep" = 16, "Dairy cattle" = 17,
                                "Beef cattle" = 15), name = NULL) +
  scale_y_continuous(breaks = seq_along(top5_regions), labels = rev(top5_regions)) +
  scale_x_continuous(labels = label_comma(accuracy = 0.1),
                     expand = expansion(mult = 0.07)) +
  labs(
    title    = sprintf("Change in head counts, %d–%d, five largest sheep declines",
                       START_YEAR, END_YEAR),
    subtitle = "Millions of head. Head counts are not feed-equivalent",
    x = "Million head, change over the period", y = NULL,
    caption  = CAPTION
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/02-substitution.png", p2,
       width = 9, height = 5.5, dpi = 200, bg = "white")
p2
```

Stated plainly, per hundred sheep lost:

```{r}
#| label: tbl-substitution
#| tbl-cap: "Cattle gained (+) or lost (−) per 100 sheep lost, five largest sheep declines, ordered by sheep lost."
substitution |>
  filter(region %in% top5_regions) |>
  arrange(Sheep) |>
  transmute(
    Region = region,
    `Sheep lost` = fmt(-Sheep),
    `Dairy cattle per 100 sheep lost` = sprintf("%+.1f", per_100(region, "Dairy cattle")),
    `Beef cattle per 100 sheep lost`  = sprintf("%+.1f", per_100(region, "Beef cattle"))
  ) |>
  kable(align = "lrrr")
```

Canterbury accounts for by far the largest absolute dairy increase of any region,
at `r fmt(substitution$"Dairy cattle"[substitution$region == "Canterbury"]/1e3)`
thousand head. It is not, however, the highest *ratio*: West Coast gained
`r fmt(per_100("West Coast", "Dairy cattle"), 0)`
dairy cattle per 100 sheep lost, on a base so small — under
`r fmt(-substitution$Sheep[substitution$region == "West Coast"]/1e3)` thousand
sheep — that the ratio says more about the denominator than about land use.
`r and_list(lost_cattle_class)` lost sheep and at least one cattle class. A
negative figure in this table means that class of animal declined as well.

## Data quality checks

Five rules run over the analysis table on every render.

```{r}
#| label: tbl-validation
#| tbl-cap: "Validation rules on the analysis table. Items: cells checked. nNA: cells where the rule could not be evaluated."
kable(results, align = "lrrrr",
      col.names = c("Rule", "Items", "Passes", "Fails", "Not evaluable (nNA)"))
```

`r number_word(n_pass)` pass. The fifth fires `r results$fails[results$rule == "census_year_reported"]`
times, and that is a finding rather than a defect. `census_year_reported` asserts
that a census year — a full-coverage collection, with no sampling error to
exceed a threshold — should not have withheld cells. It is wrong. Census years do
withhold cells, because withholding is about confidentiality and imputation
quality, not about coverage. The rule earns its place by being falsified. It sees
only flagged cells, so a cell absent from the export is invisible to it:
`r results$fails[results$rule == "census_year_reported"]` is a lower bound on
census-year withholding, not a count of it.

Three of the four passing rules cannot fail on anything the loader returns,
because ingestion already stops on a negative count, a value without a flag or a
flag without a value, and an unmapped region code. On the real data their zeroes
restate the ingestion contract; the corrupted copy below is what shows that the
rules themselves fire.

### Showing that the rules can fail

An all-green table is not evidence. A reader cannot tell from it whether the
rules are sound or merely too loose to fire. So the same five rules are also run
against a deliberately corrupted copy of the analysis table: one negative count,
one withheld cell filled with a zero, one census-year value removed, one region
label deleted, and one duplicated row.

```{r}
#| label: tbl-validation-corrupted
#| tbl-cap: "The same rules on the real data and on a deliberately corrupted copy."
results |>
  select(rule, `real data` = fails) |>
  left_join(select(results_corrupted, rule, `corrupted copy` = fails),
            by = "rule") |>
  kable(align = "lrr", col.names = c("Rule", "Fails on real data",
                                     "Fails on corrupted copy"))
```

Each corruption moves exactly one rule. `no_duplicate_cells` moves by two
because a duplicate makes both members of the pair non-unique, and
`census_year_reported` moves from `r results$fails[results$rule == "census_year_reported"]`
to `r results_corrupted$fails[results_corrupted$rule == "census_year_reported"]`
because it already fails
`r results$fails[results$rule == "census_year_reported"]` times on the real data.
The corrupted rows are chosen at distinct indices and the function asserts as
much, so no single corruption can be counted twice.

The more interesting result is in the reconciliation: the published national
total against the sum of the regions.

```{r}
#| label: tbl-recon
#| tbl-cap: "Sheep: sum of published regions against the published national total, by year."
c_flag_sheep_years <- grid$year[grid$status == "C" & grid$livestock_class == "Sheep"]
reconciliation |>
  filter(livestock_class == "Sheep") |>
  left_join(select(coverage, year, regions_present, regions_suppressed, regions_absent), by = "year") |>
  transmute(
    Year = year,
    `Region rows present` = regions_present,
    `Region rows absent` = regions_absent,
    `Regions withheld` = regions_suppressed,
    `Sum of regions` = fmt(region_sum),
    `Published national` = fmt(published),
    `Residual (head)` = ifelse(year %in% c_flag_sheep_years,
                               "not shown (C flag)", fmt(residual)),
    `Residual %` = ifelse(year %in% c_flag_sheep_years, "",
                          ifelse(residual != 0 & abs(residual_pct) < 0.001,
                                 "<0.001", sprintf("%.3f", residual_pct)))
  ) |>
  kable(align = "rrrrrrrr")
```

There are `r N_REGIONS` expected regional rows for sheep in each year. An
**absent row** is different from a present row whose value is withheld. The
exported coverage table records both: `suppression_rate` divides withheld rows
by present rows, and `suppression_rate_all_regions` divides them by
`r N_REGIONS`. Neither treats absent rows as withheld or as zero-valued
livestock counts.

The residual column is reported in head as well as in percent on purpose. At a
base of 30 million, a percentage rounded to three decimal places renders a
discrepancy of one animal as `0.000`, which would let the table appear to
support a stronger claim than the numbers do; non-zero residuals below that
resolution are shown as `<0.001`.

A residual is arithmetic on published cells, but when a single region is
withheld it equals that region's value to within a head or two. Stats NZ does not
apply secondary suppression to the national totals, so this is a property of
the published table; the page still does not report residuals for class-years
carrying the confidentiality flag `C`, or describe any residual as a named
region's value. `outputs/reconciliation.csv` carries every residual, because
the port check compares it.

Across all three livestock classes the residual separates into three tiers.

```{r}
#| label: tbl-tiers
#| tbl-cap: "Every class-year, classified by whether its region row is complete and how closely it reconciles."
residual_tiers |>
  count(tier, name = "Class-years") |>
  left_join(
    residual_tiers |>
      group_by(tier) |>
      summarise(`Largest absolute residual (head)` = fmt(max(abs(residual))),
                .groups = "drop"),
    by = "tier"
  ) |>
  rename(Tier = tier) |>
  kable(align = "lrr")
```

In `r sum(residual_tiers$tier == "Exact")` of the
`r sum(residual_tiers$fully_published)` class-years where all `r N_REGIONS`
regions are published and nothing is withheld, the regions sum to the published
national total **exactly, to the head**.

Completeness is checked before the residual, so a class-year with an incomplete
region row is never Exact however well it reconciles. That matters here:
`r sum(residual_tiers$residual == 0)` class-years reconcile to zero,
`r sum(residual_tiers$residual == 0) - sum(residual_tiers$tier == "Exact")`
more than the Exact tier contains, because
`r sum(residual_tiers$residual == 0 & residual_tiers$tier != "Exact")`
class-year with an incomplete region row happens to reconcile exactly anyway.

In `r nrow(off_tier)` further fully-published class-years the residual never
exceeds `r max(abs(off_tier$residual))` head, on bases between
`r fmt(min(off_tier$published)/1e6, 1)` and
`r fmt(max(off_tier$published)/1e6, 1)` million. I cannot attribute that from
published data. It is consistent with the national aggregate being rounded
independently of the regional cells, but I have not been able to source that
and do not assert it. It is not input perturbation: every one of these
class-years falls between `r min(off_tier$year)` and `r max(off_tier$year)`,
before perturbation was introduced in 2017, and the
`r nrow(fp_recent)` fully-published class-years from 2017 on have residuals of
`r and_list(unique(fmt(fp_recent$residual)))`.

Every residual larger than that — and the next smallest is
`r fmt(next_smallest)` head — belongs to a year whose region row is incomplete,
because a region is withheld or because a region row is absent from that year's
export.

What the one-head residuals are *not* is an artefact of this analysis. The same
discrepancy appears between cells that this analysis never sums. Adding the two
published island totals and comparing them against the published national total —
three aggregate cells, always published, never withheld — gives a difference of
one head in
`r sum(island_reconciliation$residual != 0)` of the
`r nrow(island_reconciliation)` class-years, and never more than
`r max(abs(island_reconciliation$residual))`.

```{r}
#| label: tbl-island
#| tbl-cap: "Class-years in which the two published island totals do not sum to the published national total."
island_reconciliation |>
  filter(residual != 0) |>
  transmute(
    Year = year,
    Class = livestock_class,
    `North Island` = fmt(a10),
    `South Island` = fmt(a19),
    `Sum` = fmt(a10 + a19),
    `Published NZ` = fmt(a20),
    `Difference` = fmt(residual)
  ) |>
  kable(align = "rlrrrrr")
```

So the one-head discrepancies are a property of the published table, not of the
join. That does not explain them, but it does locate them.

### Withholding is not constant over time

```{r}
#| label: fig-suppression
#| fig-cap: !expr 'sprintf("Share of the %d expected region-by-class cells each year that are withheld with a flag, or absent from the export.", cells_per_year)'
#| fig-alt: !expr 'sprintf("Stacked bar chart of the percentage of the %d expected regional cells each year, %d to %d, that are not published, split into three kinds. From %d to %d no cell carries a flag, but between %d and %d cells a year are absent from the export. In %s the C confidentiality flag appears on %d cells. The S quality-suppression flag appears in %s, then every year from %d to %d, peaking at %s percent in %d. Census years are shaded.", cells_per_year, START_YEAR, END_YEAR, min(absent_years), max(absent_years), min(status_by_year$n[status_by_year$status == "Absent"]), max(status_by_year$n[status_by_year$status == "Absent"]), and_list(c_years), sum(grid$status == "C"), and_list(s_years[s_years < 2017]), 2017, END_YEAR, fmt(peak_row$rate, 1), peak_row$year)'
status_labels <- c("C" = "C flag: confidentiality (before 2017)",
                   "S" = "S flag: quality suppression",
                   "Absent" = "Absent from the export: no row")
p3 <- status_by_year |>
  filter(status != "Published") |>
  mutate(status = factor(unname(status_labels[status]), levels = rev(unname(status_labels)))) |>
  ggplot(aes(x = year, y = rate)) +
  annotate("rect", xmin = CENSUS_YEARS - 0.5, xmax = CENSUS_YEARS + 0.5,
           ymin = -Inf, ymax = Inf, fill = "grey93") +
  geom_col(aes(fill = status), width = 0.7) +
  scale_fill_manual(values = setNames(c("#e07a5f", "#6c8ea4", "#c9c3b6"),
                                      unname(status_labels)), name = NULL,
                    breaks = unname(status_labels)) +
  scale_x_continuous(breaks = sort(unique(c(START_YEAR, CENSUS_YEARS, END_YEAR))),
                     limits = c(START_YEAR - 1, END_YEAR + 1)) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  guides(fill = guide_legend(nrow = 2)) +
  labs(
    title    = "Rows are absent until 2011; confidentiality flags appear in 2012; quality suppression from 2014, peaking in 2021",
    subtitle = sprintf("Percentage of the %d expected region-by-class cells not published: sheep, dairy and beef cattle. Shaded: census years",
                       cells_per_year),
    x = NULL, y = "Cells not published",
    caption  = CAPTION
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold", size = 11),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/03-suppression-rate.png", p3,
       width = 9, height = 5, dpi = 200, bg = "white")
p3
```

Reading that chart as a single rising trend would be wrong, and the `OBS_STATUS`
column is what makes the difference visible. Two flags are in use, they are not
the same mechanism, and before either appears the export does something else
entirely.

**Before `r first_flag_year`, cells are left out rather than flagged.** From
`r min(absent_years)` to `r max(absent_years)` no cell carries a flag, but the
export has fewer rows than the full grid of regions and classes; from
`r first_flag_year` on it is complete. An absent row is not a zero. In
`r gap_nonzero` of the `r nrow(unflagged_gaps)` class-years whose region row is
incomplete without any flag, the published regions fall short of the published
national total, by between `r fmt(min(unflagged_gaps$residual[unflagged_gaps$residual != 0]))`
and `r fmt(max(unflagged_gaps$residual))` head, so the animals were counted and
not shown`r if (gap_nonzero < nrow(unflagged_gaps)) "; the exception reconciles exactly, as it would if the missing region had none of that class" else ""`.
And the cells that carry `C` in 2012 look like the cells that were absent
before: across all 44 livestock codes, `r sum(c_2012$absent >= 1)` of the
`r nrow(c_2012)` regional `C` cells of 2012 had been absent from the export in
at least one earlier year, on average `r fmt(mean(c_2012$absent), 1)` of the
`r length(START_YEAR:2011)` years from `r START_YEAR` to 2011. The change in
`r first_flag_year` is a change in the export's format; it is not evidence that
nothing was withheld before it, and this chart is not a trend in withholding
before `r first_flag_year`.

The `C` flag is confidentiality suppression. Among the three livestock classes
used here it appears in `r and_list(c_years)` and in no other year.
Checked across every one of the 44 livestock codes in the raw extract, it appears
in `r paste(range(all_c_years), collapse = "–")` and **never after
`r max(all_c_years)`**.

That is what the published methodology predicts. Stats NZ states that prior to
2017, figures affected by the confidentiality provisions were denoted by `C`,
and that since 2017 confidentiality has instead been implemented by an input
perturbation method that adds noise at individual farm level, so output figures
no longer have to be replaced with `C`. The flag stops in the file in the same
year the method changed.

The `S` flag is quality suppression, applied where sampling errors or imputation
levels are high. It appears in `r and_list(s_years[s_years < 2017])`, then every
year from 2017 to `r END_YEAR`, peaking at `r fmt(peak_row$rate, 1)` percent of
cells in `r peak_row$year`. The census years hold few `S` cells —
`r s_count(2017)` in 2017 and `r s_count(2022)` in 2022, against
`r s_count(2024)` in the survey year 2024 — which is consistent with a census
removing the sampling-error branch of the rule while leaving the imputation
branch. Two census years are far too few to establish it. All
`r nrow(s_census)` census-year `S` cells in this table are
`r and_list(s_census_regions)`.

So the 2012 spike is not a quality problem. All
`r sum(grid$year == 2012 & grid$status %in% c("C", "S"))` withheld cells that
year carry `C`, not `S`.

One of the two regions this analysis has to drop is explained in the published
methodology rather than guessed at. Sampling error could not be calculated for
Nelson because only one responding unit was observed in each sampled stratum,
and at least two are needed to estimate a within-stratum variance. Auckland, the
other dropped region, is simply withheld, with no reason published.

## Method

1. **One table, one vintage.** Stats NZ table `AGR_AGR_003`, *Livestock Numbers
   by Regional Council*, dataflow `STATSNZ:AGR_AGR_003(1.0)`, final vintage,
   downloaded once and committed with its SHA-256. Full provenance is in
   [`data-raw/SOURCE.md`](https://github.com/Kenchch/nz-sheep-decline-by-region/blob/main/data-raw/SOURCE.md).
2. **Codes verified before code was written.** The extract carries numeric codes
   and no labels. Sheep, dairy cattle and beef cattle were each confirmed against
   the published June 2024 national figures before any analysis was run. The
   `LIVESTOCK` codelist holds 44 entries that nest inside one another; adding all
   44 national totals for 2024 gives `r fmt(all_codes_2024/1e6, 1)` million,
   against `r fmt(three_codes_2024/1e6, 1)` million for the three verified totals
   alone. Only those three codes are used.
3. **Aggregates are separated, not summed.** `AREA` codes 10, 19 and 20 are the
   North Island, South Island and New Zealand totals. They are flagged in the
   analysis table and excluded from every regional sum. The extract publishes
   no area labels at all, only numeric codes that are this table's own rather
   than SSGA23 codes; region names here follow the Statistical standard for
   geographic areas 2023 (SSGA23), including macrons, which differ from the
   older spellings the codelist still displays in Aotearoa Data Explorer.
4. **Withheld cells stay withheld.** A withheld cell becomes `NA` and carries a
   flag. It is never filled with zero and never dropped. Regions missing either
   endpoint are excluded from the change calculation and counted in the text.
5. **Reconciliation is reported, not asserted.** No fixed tolerance is set
   between the regional sum and the published total. The residual is computed for
   every year and published above.
6. **Quantitative claims are computed or cited, and the two are different
   things.** Numbers that come from the data — in the prose, the figure and
   table captions and the figures' alternative text — are expressions evaluated
   at render time against the analysis table. Qualitative claims that depend on
   the data (for example, that every measurable region declined) are asserted in
   the code, so a new vintage that breaks one stops the render instead of
   leaving the sentence stale. Numbers from Stats NZ's published methodology —
   sample size, eligible population, response rate, imputation level, relative
   sampling error, the GST threshold — *cannot* be computed from an aggregate
   extract; they are typed, and each is attributed under [Sources for the
   methodological statements](#sources-for-the-methodological-statements). The
   analysis window, the census years, the `r DAIRY_PEAK_YEAR` split and the
   five-year minimum window are design constants, set once in `R/load.R`,
   `R/checks.R` and the setup code of this page.

### Release schedule

This analysis uses the final vintage throughout. Up to and including the 2024
survey, Stats NZ published a provisional release in mid-December and a final
release in mid-May of the following year. From the 2025 survey those two were
replaced by a single final release in mid-April.

### Survey design

The reference period is the year ended 30 June. Censuses were held in
`r and_list(CENSUS_YEARS)`, with annual sample surveys in the intervening years.
The sample is stratified by regional council area, ANZSIC06 group and size group.
The 2025 survey drew 24,600 geographic locations from an estimated eligible
population of 45,800; note that a geographic location is the statistical unit
and is not the same thing as a farm.

The series starts at `r START_YEAR` because Stats NZ states that the population
for the 2002 census and subsequent surveys differs from that of earlier
agricultural censuses and surveys, and that figures from 2002 onward may not be
directly comparable with previous years. That is also why the 1994 observation
carried in the same table is excluded here. There was no agricultural
production survey at all in 1997, 1998 or 2001.

Values for non-respondents are imputed by random hot deck imputation, with the
imputation class formed from regional council area, ANZSIC06 group and production
data from previous years. This matters more than it may appear: for the 2025
survey Stats NZ reports that **30 percent** of the total sheep estimate was
imputed, against a relative sampling error of 3 percent at 95 percent confidence.
The response rate for the 2025 final release was 72.1 percent. The collection is
run by Stats NZ in partnership with the Ministry for Primary Industries.

#### The headline window is not design-matched

`r START_YEAR` is a census and `r END_YEAR` is a sample survey, so the headline
differences a full-coverage count against an estimate that carries both sampling
error and a 30 percent imputation rate. Stats NZ publishes both years in the
same table, but the comparison is still an asymmetry, and this report carries
the `is_census_year` flag through the analysis table precisely so that it can be
checked rather than assumed.

The census-to-census window that avoids the asymmetry is `r START_YEAR` to
`r LAST_CENSUS`:

```{r}
#| label: tbl-matched
#| tbl-cap: "The headline window against the design-matched census-to-census window."
rbind(
  data.frame(Window = sprintf("%d–%d, census to sample survey (headline)", START_YEAR, END_YEAR),
             Regions = headline$regions,
             Fall = fmt(headline$fall),
             `Top three share` = sprintf("%.1f%%", headline$share),
             `Their share of the flock` = sprintf("%.1f%%", headline$base_share),
             `The three regions` = headline$top3,
             check.names = FALSE),
  data.frame(Window = sprintf("%d–%d, census to census (design-matched)", START_YEAR, LAST_CENSUS),
             Regions = matched$regions,
             Fall = fmt(matched$fall),
             `Top three share` = sprintf("%.1f%%", matched$share),
             `Their share of the flock` = sprintf("%.1f%%", matched$base_share),
             `The three regions` = matched$top3,
             check.names = FALSE)
) |> kable(align = "lrrrrl")
```

The design-matched window is measurable across `r matched$regions - headline$regions`
more region`r if (matched$regions - headline$regions != 1) "s" else ""`, because
`r and_list(matched_gain)` is published in `r LAST_CENSUS` and withheld in
`r END_YEAR`. The share is similar:
a fall of `r fmt(matched$fall/1e6, 1)` million head with the three largest
decliners accounting for `r fmt(matched$share, 1)` percent, against
`r fmt(matched$base_share, 1)` percent of the flock. The three are not the same:
`r and_list(added_in_matched)` replaces `r and_list(dropped_in_matched)`. The
share is robust; the list is not.

The `r END_YEAR` endpoint is used in the headline because it is the most recent
published figure, not because it is the better-matched one.

### What was excluded, and why

- **Territorial authority geography** — the surveys are designed to produce
  results to regional council level and the censuses to territorial authority
  level, so no annual series exists below the region.
- **Horticulture** — not collected in 2004, 2006, 2008, 2010, 2013, 2015, 2016,
  2018, 2021, 2023 or 2025, which is most of this window.
- **Forestry production** — Stats NZ no longer collects or compiles forestry
  production statistics; they are published by MPI. Stats NZ does still collect
  forestry *land use*. The 2025 release also notes that additional National
  Exotic Forest Description data has been received and that revisions to earlier
  land use and forestry statistics, especially for 2016 to 2021, are planned.
- **Pre-2002 data** — the population change described above, plus the absence of
  any survey in 1997, 1998 and 2001.
- **A choropleth map** — a sorted bar chart answers "which regions contributed
  most" and a map does not, because a map encodes land area rather than
  magnitude. Canterbury, Otago and Southland are also three of the largest
  regions by area, which is precisely the confound a map would introduce.
- **A second data set** — the question is answerable inside one table, and every
  extra source adds provenance to defend without adding evidence.

### Implications for output design

Three observations follow from the checks above, and they concern the shape of
the output rather than the sheep.

Regional council is the only geography available annually; territorial authority
exists only in census years, so any sub-regional time series is structurally
unavailable rather than merely unpublished. Withholding has exceeded ten percent
of cells in `r recent$above` of the last `r recent$of` years, and it is not
spread evenly. Nelson and the Chatham Islands account for
`r fmt(100 * sum(regions$suppressed & regions$region %in% c("Nelson", "Chatham Islands")) / sum(regions$suppressed), 0)`
percent of all withheld cells in this table between them, Auckland for a
further `r fmt(100 * sum(regions$suppressed & regions$region == "Auckland") / sum(regions$suppressed), 0)`
percent, and
`r length(setdiff(unique(regions$region[regions$suppressed]), c("Nelson", "Chatham Islands", "Auckland")))`
other regions for the rest.

Two different grounds are at work, and it is worth separating them. Nelson and
the Chathams are withheld persistently — they hold the two smallest livestock
populations in the table — which makes their *trends* unusable. (By sheep alone,
averaged over the years each is published, the second smallest is
`r sheep_mean$region[2]`, published in `r sheep_mean$years[2]` of the
`r length(EXPECTED_YEARS)` years: being small and being withheld are not the
same thing.) Auckland is withheld only at scattered points including the
`r END_YEAR` endpoint, which is what removes it from this analysis' *change*
calculation while leaving most of its series intact. Auckland is not a small
region, and by sheep it is not even among the smallest: its last published count
is larger than Northland's or West Coast's. Being withheld and being small are
related here but not the same thing, and an output that conflated them would
mislead.

A suppression-aware grouping that pooled the persistently withheld regions would
preserve more signal than withholding them one at a time does. And the
alternating census-and-survey design makes every regional series in this table
structurally uneven, which belongs on the output rather than being left for a
user to discover from a reconciliation residual.

There is also a class of question this data cannot answer at all: anything at
farm level, or any breakdown finer than what is published. The access route for
those questions is approved microdata under the Five Safes framework, which I
have not used here.

### Sources for the methodological statements

Every statement about the collection above was read on the page cited here, not
carried over from secondary summaries. Pages read 4 September 2026.

| Statement | Source |
|---|---|
| Census years 2002, 2007, 2012, 2017, 2022, with sample surveys in between | [Agricultural Production Statistics: June 2025 (Final), DataInfo+](https://datainfoplus.stats.govt.nz/item/nz.govt.stats/a9908d00-4827-46b2-a38a-310c0c75b029) |
| Eligible population 45,800; sample 24,600 geographic locations | same page |
| Stratification by regional council, ANZSIC06 group and size group | same page |
| 2025 response rate 72.1 percent | same page |
| 30 percent of the 2025 sheep estimate imputed; 3 percent relative sampling error at 95 percent confidence | same page |
| Random hot deck imputation and the imputation class | same page |
| GST registration level of \$60,000; coverage below it "partial and unquantifiable" | same page |
| `C` before 2017; input perturbation since | same page |
| `S` for data with high sample errors or imputation levels | same page |
| Nelson sampling error not estimable (one responding unit per stratum) | same page |
| Horticulture collection years; forestry production now published by MPI | same page |
| Programme run in partnership with MPI | [Agriculture Production Statistics series, DataInfo+](https://datainfoplus.stats.govt.nz/item/nz.govt.stats/6362a469-f374-412e-ac25-d76fd0962003) |
| Provisional mid-December plus final mid-May up to 2024; single final mid-April from 2025 | same page |
| 2002 population differs from earlier collections; no survey in 1997, 1998, 2001 | same page |
| Regional council annually; territorial authority in censuses | same page |
| Release date of 16 April 2026 for the year to June 2025; NEFD data and planned revisions for 2016–2021 | [Agricultural production statistics: Year to June 2025](https://www.stats.govt.nz/information-releases/agricultural-production-statistics-year-to-june-2025/) |
| Mana Ōrite Relationship Agreement signed with the Data Iwi Leaders Group of the National Iwi Chairs Forum on 30 October 2019 | [Mana Ōrite Relationship Agreement](https://www.stats.govt.nz/about-us/what-we-do/mana-orite-relationship-agreement/) |

Statements that could not be sourced on those pages are not made anywhere in
this report.

## Limitations

- Non-census years are **sample estimates**. I have not recomputed their sampling
  errors and I have applied no weighting of my own. A year-on-year movement in a
  single small region should not be read as real change, and regional rankings
  separated by small margins — such as `r top3$region[3]` and
  `r fourth$region` — should not be read as ranked at all.
- Withheld cells are retained as `NA` and counted. Two regions are excluded
  from the `r START_YEAR`–`r END_YEAR` change entirely for that reason.
- The sheep-to-cattle comparisons are **ratios of head counts**. They are not
  claims about stocking rate, feed demand, land use or emissions.
- The frame is Stats NZ's Business Register, which is built on GST registration.
  The compulsory registration level is \$60,000, and Stats NZ describes coverage
  of units below that level as "partial and unquantifiable".
- **Imputation is large.** Stats NZ reports 30 percent of the 2025 national sheep
  estimate as imputed. That national rate does not establish the imputation rate
  for any individual region or another year; this extract does not provide those
  rates, and this analysis does not adjust the published estimates. Whether
  hot-deck imputation biases changes towards zero has not been assessed here.
- Since 2017, confidentiality has been implemented by input perturbation, which
  adds noise at individual farm level. Its size is not published on the pages
  cited here, so it is not quantified; it is present in every year from 2017 on.
- **The endpoints are of different collection designs.** `r START_YEAR` is a
  census and `r END_YEAR` is a sample survey. The `is_census_year` flag is carried
  through the analysis table for exactly this reason, and the design-matched
  comparison is reported under [Survey design](#survey-design).
- **Choosing the start and end years is a choice, and the two halves of the
  headline survive it differently.** See the sensitivity analysis below.

### How much does the start year matter?

Recomputing the headline statistic from every start year that leaves a window
of at least `r MIN_WINDOW` years — `r START_YEAR` to `r LAST_START` — against the
same `r END_YEAR` endpoint. The five-year minimum is an analysis choice to focus
on longer windows. Starts in `r LAST_START + 1`–`r END_YEAR - 1` are excluded, so
the sensitivity conclusions apply only to the windows shown:

```{r}
#| label: tbl-sensitivity
#| tbl-cap: !expr 'sprintf("The headline statistic from every start year, %d to %d, against the %d endpoint.", START_YEAR, LAST_START, END_YEAR)'
sensitivity |>
  transmute(
    `Start year` = ifelse(start %in% CENSUS_YEARS,
                          paste0(start, " (census)"), as.character(start)),
    `Regions measurable` = regions,
    `Gross fall (head)` = fmt(fall),
    `Net fall (head)` = fmt(net_fall),
    `Top three share` = sprintf("%.1f%%", share),
    `Their share of the flock` = sprintf("%.1f%%", base_share),
    `The three regions` = top3
  ) |>
  kable(align = "lrrrrrl")
```

`Gross fall` sums losses only in the declining regions with both endpoints
published; it is the denominator for `Top three share`. `Net fall` subtracts
gains in other measurable regions. Neither is the published national fall,
because some regional endpoints are withheld or absent. For
`r START_YEAR`–`r END_YEAR` every measurable region declined, so the gross and
net regional falls are both `r fmt(headline$fall)` head against a published
national fall of `r fmt(nat_drop)`, a gap of `r fmt(national_gap_pct, 2)`
percent, as set out at the top of this page.

Across the `r nrow(sensitivity)` start years the three largest-declining regions
account for between `r fmt(min(sensitivity$share), 0)` and
`r fmt(max(sensitivity$share), 0)` percent of the gross regional fall, while
holding between `r fmt(min(sensitivity$base_share), 0)` and
`r fmt(max(sensitivity$base_share), 0)` percent of the starting flock. The
narrower claim "over half" does not hold for every start year: the share is
below 50 percent for `r and_list(sensitivity$start[sensitivity$share < 50])`.

Which three regions they are is not robust. Manawatū-Whanganui is in the top
three in `r count_in_top3(sensitivity$top3, "Manawatū-Whanganui")` of the
`r nrow(sensitivity)` windows, Otago in `r count_in_top3(sensitivity$top3, "Otago")`,
Southland in `r count_in_top3(sensitivity$top3, "Southland")`, and Canterbury in
only `r count_in_top3(sensitivity$top3, "Canterbury")` —
`r and_list(sensitivity$start[vapply(strsplit(sensitivity$top3, ", ", fixed = TRUE), function(x) "Canterbury" %in% x, logical(1))])`.
Canterbury dominates the window this report uses because its decline is
concentrated in the early years; from a start year of 2009 or later it usually
drops out of the top three altogether.

That is worth stating plainly, because Canterbury is the region this report
names most often. It leads the `r START_YEAR`–`r END_YEAR` window, and it is the
one region where the substitution story carries weight. It is not a permanent
feature of the data.

### And the end year?

Holding the start at `r START_YEAR` and moving the end instead, from
`r FIRST_END` to `r END_YEAR`. The measurable regions change from window to
window, so the shares do not share a denominator; the last column names the
regions each window excludes.

```{r}
#| label: tbl-end-sensitivity
#| tbl-cap: !expr 'sprintf("The headline statistic from %d to every end year, %d to %d.", START_YEAR, FIRST_END, END_YEAR)'
end_sensitivity |>
  transmute(
    `End year` = ifelse(end %in% CENSUS_YEARS,
                        paste0(end, " (census)"), as.character(end)),
    `Regions measurable` = regions,
    `Top three share` = sprintf("%.1f%%", share),
    `Their share of the flock` = sprintf("%.1f%%", base_share),
    `The three regions` = top3,
    `Not measurable` = ifelse(excluded == "", "—", excluded)
  ) |>
  kable(align = "lrrrll")
```

From every end year the three largest declines take more of the fall than they
held of the flock, by between
`r fmt(min(end_sensitivity$share - end_sensitivity$base_share), 1)` and
`r fmt(max(end_sensitivity$share - end_sensitivity$base_share), 1)` percentage
points. Canterbury and Southland are in the top three for every end year; the
third place moves between
`r and_list(sort(unique(vapply(strsplit(end_sensitivity$top3, ", ", fixed = TRUE), function(x) x[3], character(1)))))`.

```{r}
#| include: false
stopifnot("the end-year prose names Canterbury and Southland as ever-present" =
            count_in_top3(end_sensitivity$top3, "Canterbury") == nrow(end_sensitivity),
          count_in_top3(end_sensitivity$top3, "Southland") == nrow(end_sensitivity),
          all(end_sensitivity$share > end_sensitivity$base_share))
```

## Ethics

This analysis uses published regional aggregates only. No microdata and no IDI
access is involved, so the Five Safes framework does not apply to it.

Stats NZ lists Tatauranga Umanga Māori among the internal users of this
collection, and signed the Mana Ōrite Relationship Agreement with the Data Iwi
Leaders Group of the National Iwi Chairs Forum in October 2019. Any Māori-specific
cut of this data therefore sits inside an established relationship between Stats
NZ and iwi Māori. It does not sit with an unaffiliated analyst working from
published aggregates, which is why this report does not attempt one and does not
characterise the terms of that agreement.

## Reproduce

The full command sequence — restoring the pinned packages, the ingestion and
helper tests, regenerating the CSVs and rendering this page — is in the
[README](https://github.com/Kenchch/nz-sheep-decline-by-region#reproduce).

`R/load.R` reads the SHA-256 recorded in `data-raw/SOURCE.md` and stops unless
the extract's bytes match it, then tidies the extract; `R/checks.R` runs the
rules and both reconciliations; `index.qmd` produces this page, the figures and
`outputs/key-figures.csv`.

Both sets of outputs are committed. A clean checkout that runs the documented
commands must reproduce the CSVs byte for byte. The figures are regenerated as
well, but their bytes can differ with the platform's fonts and graphics device,
so they are checked for existence rather than compared.

That claim is checked rather than asserted. A GitHub Actions workflow runs the
commands on a clean Ubuntu runner for every pull request and every push to
`main`, and weekly, and fails if any committed CSV differs. On `main`, the site
it rendered and verified is the one deployed to GitHub Pages, so what you are
reading is the output of a build that passed.

It earned its place on the first run, by failing. The extract arrives from the
SDMX endpoint with CRLF line endings and the recorded SHA-256 is the hash of
those bytes, but Git had normalised them to LF on commit. The hash gate
therefore passed on the Windows machine this was written on, where checkout
restores the CRLFs, and would have stopped every clone on Linux or macOS at the
first line of `R/load.R`. `.gitattributes` now marks `data-raw/*.csv` as `-text`
so Git leaves those bytes alone.

The general point is worth more than the fix: a single machine cannot see a
whole class of defect, because the thing that hides the defect is the same thing
that makes the analysis run. Only a second platform could find that one.

::: {.callout-note collapse="true"}
## Session information

```{r}
#| label: session
#| echo: false
sessionInfo()
```
:::
