---
title: "Where New Zealand's sheep went"
subtitle: "A regional breakdown of the flock decline, 2002–2025"
author: "Feng Jiang"
date: today
---

```{r setup}
#| include: false
source("R/load.R")
source("R/checks.R")
suppressPackageStartupMessages({
  library(ggplot2); library(scales); library(tidyr); library(knitr)
})

dir.create("outputs/figures", recursive = TRUE, showWarnings = FALSE)

d       <- livestock
regions <- regions_only(d)

fmt <- function(x, digits = 0) format(round(x, digits), big.mark = ",", trim = TRUE)

# Sheep, first year to last year, by region -------------------------------
sheep_change <- regions |>
  filter(livestock_class == "Sheep", year %in% c(2002, 2025)) |>
  select(region, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002, pct = 100 * change / y2002) |>
  arrange(change)

measurable  <- sum(!is.na(sheep_change$change))
decline_sum <- sum(sheep_change$change[sheep_change$change < 0], na.rm = TRUE)
top3        <- head(sheep_change, 3)
top3_share  <- 100 * sum(top3$change) / decline_sum

nat_sheep <- national(d) |> filter(livestock_class == "Sheep")
nat_2002  <- nat_sheep$head[nat_sheep$year == 2002]
nat_2025  <- nat_sheep$head[nat_sheep$year == 2025]
nat_drop  <- nat_2002 - nat_2025
nat_pct   <- 100 * nat_drop / nat_2002

nat_wide <- national(d) |>
  filter(year %in% c(2002, 2025)) |>
  select(livestock_class, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002)

dairy_gain <- nat_wide$change[nat_wide$livestock_class == "Dairy cattle"]
beef_gain  <- nat_wide$change[nat_wide$livestock_class == "Beef cattle"]

# Substitution, all three classes ----------------------------------------
substitution <- regions |>
  filter(year %in% c(2002, 2025)) |>
  select(region, livestock_class, year, head) |>
  pivot_wider(names_from = year, values_from = head, names_prefix = "y") |>
  mutate(change = y2025 - y2002) |>
  select(region, livestock_class, change) |>
  pivot_wider(names_from = livestock_class, values_from = change)

top5_regions <- head(sheep_change$region, 5)

dairy_both    <- substitution |> filter(!is.na(Sheep), !is.na(`Dairy cattle`))
n_dairy_falls <- sum(dairy_both$`Dairy cattle` < 0)
n_dairy_both  <- nrow(dairy_both)

# Suppression rate --------------------------------------------------------
supp <- regions |>
  group_by(year, is_census_year) |>
  summarise(cells = n(), suppressed_cells = sum(suppressed), .groups = "drop") |>
  mutate(rate = 100 * suppressed_cells / cells)
```

::: {.callout-note appearance="simple"}
Source data: Stats NZ, *Agricultural production statistics*, licensed by Stats NZ
for re-use under the Creative Commons Attribution 4.0 International licence.
:::

## The short version

New Zealand's sheep flock fell by **`r fmt(nat_drop/1e6, 2)` million** between June
2002 and June 2025, a drop of `r fmt(nat_pct, 0)` percent.

Three points, each with one number.

- Three regions — `r paste(top3$region, collapse = ", ")` — account for
  **`r fmt(top3_share, 0)` percent** of the total fall, across the
  `r measurable` regions where both endpoints are published.
- Over the same 23 years the national dairy herd grew by only
  **`r fmt(dairy_gain/1e3, 0)` thousand** cattle, while beef cattle *fell* by
  `r fmt(-beef_gain/1e3, 0)` thousand.
- In `r n_dairy_falls` of the `r n_dairy_both` regions where both series are
  published, dairy cattle fell **alongside** sheep rather than in place of them.

### What this means

The common story is that sheep farms became dairy farms. At a national level the
arithmetic does not support that story. Sheep numbers fell by roughly 16 million
head. Dairy cattle rose by roughly six hundred thousand. Those are different
orders of magnitude.

The picture changes region by region. Canterbury and Southland did add dairy
cattle on a scale worth noticing. Most other regions that lost sheep did not, and
several lost dairy cattle as well. Whatever replaced the sheep, in most of the
country it was not another animal counted in this table.

### What these numbers do not tell you

They are head counts. They say nothing about land area, feed demand, grazing
pressure, farm profitability, or what the land is used for now. A dairy cow and a
sheep are not exchangeable units, and this page does not treat them as such.

## The regional breakdown

```{r}
#| label: fig-contribution
#| fig-cap: "Contribution of each region to the national fall in sheep numbers, June 2002 to June 2025."
plot_dat <- sheep_change |>
  filter(!is.na(change)) |>
  mutate(region = factor(region, levels = rev(region)),
         millions = -change / 1e6)

p1 <- ggplot(plot_dat, aes(x = millions, y = region)) +
  geom_col(fill = "#2d6a4f", width = 0.72) +
  geom_text(aes(label = sprintf("%.2f", millions)),
            hjust = -0.15, size = 3.1, colour = "grey25") +
  annotate("label", x = max(plot_dat$millions) * 0.45,
           y = 4.2,
           label = sprintf("Canterbury, Southland and Otago\naccount for %.0f%% of the fall",
                           top3_share),
           hjust = 0, size = 3.4, label.size = 0, fill = "grey96", colour = "grey20") +
  scale_x_continuous(labels = label_comma(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.14))) +
  labs(
    title    = "Three South Island regions account for over half the national fall",
    subtitle = "Reduction in sheep numbers, millions of head, June 2002 to June 2025",
    x = "Million head fewer", y = NULL,
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        plot.title.position = "plot",
        plot.title    = element_text(face = "bold"),
        plot.caption  = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/01-regional-contribution.png", p1,
       width = 9, height = 6, dpi = 200, bg = "white")
p1
```

Two regions are missing from that chart. Auckland and Nelson have a suppressed
2025 value, so no change can be computed for them. They are excluded and counted,
never imputed and never set to zero.

## Did dairy replace the sheep?

```{r}
#| label: fig-substitution
#| fig-cap: "Change in sheep, dairy cattle and beef cattle in the five regions that lost the most sheep."
dumb <- substitution |>
  filter(region %in% top5_regions) |>
  pivot_longer(-region, names_to = "livestock_class", values_to = "change") |>
  mutate(
    region = factor(region, levels = rev(top5_regions)),
    livestock_class = factor(livestock_class,
                             levels = c("Sheep", "Dairy cattle", "Beef cattle")),
    thousands = change / 1e3
  )

p2 <- ggplot(dumb, aes(x = thousands, y = region)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_line(aes(group = region), colour = "grey75", linewidth = 0.8) +
  geom_point(aes(colour = livestock_class), size = 3.4) +
  scale_colour_manual(values = c("Sheep" = "#2d6a4f",
                                 "Dairy cattle" = "#c1666b",
                                 "Beef cattle" = "#4a6fa5"), name = NULL) +
  scale_x_continuous(labels = label_comma(), expand = expansion(mult = 0.07)) +
  labs(
    title    = "Where sheep left, cattle mostly did not arrive in comparable numbers",
    subtitle = "Change in animals, thousands of head, June 2002 to June 2025, five largest sheep declines",
    x = "Thousand head, change over the period", y = NULL,
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/02-substitution.png", p2,
       width = 9, height = 5.5, dpi = 200, bg = "white")
p2
```

Stated plainly, per hundred sheep lost:

```{r}
#| label: tbl-substitution
substitution |>
  filter(region %in% top5_regions) |>
  transmute(
    Region = region,
    `Sheep lost` = fmt(-Sheep),
    `Dairy cattle per 100 sheep lost` = sprintf("%+.1f", -100 * `Dairy cattle` / Sheep),
    `Beef cattle per 100 sheep lost`  = sprintf("%+.1f", -100 * `Beef cattle` / Sheep)
  ) |>
  kable(align = "lrrr")
```

Canterbury is the one region where the substitution story carries real weight.
Hawkes Bay and Manawatu-Wanganui lost sheep *and* cattle. A negative figure in
this table means that class of animal declined as well.

## Data quality checks

Five rules run over the analysis table on every render.

```{r}
#| label: tbl-validation
kable(results, align = "lrrrr")
```

The interesting result is not in that table. It is in the reconciliation: the
published national total against the sum of the regions.

```{r}
#| label: tbl-recon
reconciliation |>
  filter(livestock_class == "Sheep") |>
  left_join(select(coverage, year, regions_present, regions_suppressed), by = "year") |>
  transmute(
    Year = year,
    `Regions published` = regions_present,
    `Regions suppressed` = regions_suppressed,
    `Sum of regions` = fmt(region_sum),
    `Published national` = fmt(published),
    `Residual %` = sprintf("%.3f", residual_pct)
  ) |>
  kable(align = "rrrrrr")
```

The residual is zero, to the head, in every year where all seventeen regions are
published and no cell is suppressed. It becomes non-zero for exactly two reasons,
and only those two: a region is suppressed, or a region code is absent from the
table for that year. It is not a join error and it is not perturbation.

### Suppression is not constant over time

```{r}
#| label: fig-suppression
#| fig-cap: "Share of regional cells suppressed, by year."
p3 <- ggplot(supp, aes(x = year, y = rate)) +
  geom_col(aes(fill = is_census_year), width = 0.7) +
  scale_fill_manual(values = c("FALSE" = "#95a5a6", "TRUE" = "#e07a5f"),
                    labels = c("Sample survey", "Census"), name = NULL) +
  scale_x_continuous(breaks = seq(2002, 2025, 3)) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title    = "Suppression appears in 2012 and rises through the recent years",
    subtitle = "Percentage of region-by-livestock-class cells suppressed: sheep, dairy and beef cattle",
    x = NULL, y = "Cells suppressed",
    caption  = "Source: Stats NZ, Agricultural production statistics.\nLicensed by Stats NZ for re-use under CC BY 4.0."
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "top",
        plot.title.position = "plot",
        plot.title   = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey45", hjust = 0, size = 8))

ggsave("outputs/figures/03-suppression-rate.png", p3,
       width = 9, height = 5, dpi = 200, bg = "white")
p3
```

No cell is suppressed between 2002 and 2011. Suppression appears from 2012 and
rises through the later years, peaking at
`r sprintf("%.1f", max(supp$rate))` percent in `r supp$year[which.max(supp$rate)]`.
The census years 2017 and 2022 sit visibly below the survey years around them,
which is what a rule based on sampling error or imputation share would predict.
2012 is the exception, and this analysis does not explain it.

One qualification, which matters. I cannot tell from this extract whether the
zeroes before 2012 mean that no suppression was applied, or that the flag was not
carried in this export. The chart shows what the file contains.

### Independent recomputation

Stats NZ publishes a national sheep figure for June 2024. Recomputing it from the
regional cells in the same extract gives a different number, and the difference
is fully explained.

```{r}
#| label: tbl-recompute
r24 <- reconciliation |> filter(livestock_class == "Sheep", year == 2024)
data.frame(
  Quantity = c("Published, Total New Zealand, June 2024",
               "Recomputed as the sum of published regions",
               "Difference"),
  Head = c(fmt(r24$published), fmt(r24$region_sum), fmt(r24$residual))
) |> kable(align = "lr")
```

The difference is Auckland, whose 2024 value is suppressed. Filling it with zero
would have produced a total that looked tidier and was wrong by
`r fmt(r24$residual)` head.

## Method

1. **One table, one vintage.** Stats NZ table `AGR_AGR_003`, *Livestock Numbers
   by Regional Council*, dataflow `STATSNZ:AGR_AGR_003(1.0)`, final vintage,
   downloaded once and committed with its SHA-256. Full provenance is in
   [`data-raw/SOURCE.md`](https://github.com/Kenchch/nz-sheep-decline-by-region/blob/main/data-raw/SOURCE.md).
2. **Codes verified before code was written.** The extract carries numeric codes
   and no labels. Sheep, dairy cattle and beef cattle were each confirmed against
   the published June 2024 national figures before any analysis was run. The
   `LIVESTOCK` codelist holds 44 entries that nest inside one another; adding all
   44 for 2024 gives 115.7 million animals against a published 33.8 million. Only
   the three verified codes are used.
3. **Aggregates are separated, not summed.** `AREA` codes 10, 19 and 20 are the
   North Island, South Island and New Zealand totals. They are flagged in the
   analysis table and excluded from every regional sum.
4. **Suppressed cells stay suppressed.** A suppressed cell becomes `NA` and
   carries a flag. It is never filled with zero and never dropped. Regions
   missing either endpoint are excluded from the change calculation and counted
   in the text.
5. **Reconciliation is reported, not asserted.** No fixed tolerance is set
   between the regional sum and the published total. The residual is computed for
   every year and published above.

### Release schedule

This analysis uses the final vintage throughout. Up to and including the 2024
survey, Stats NZ published a provisional release in mid-December and a final
release in mid-May of the following year. From the 2025 survey those two were
replaced by a single final release in mid-April.

### Survey design

The reference period is the year ended 30 June. 2002, 2007, 2012, 2017 and 2022
are census years; every other year in the series is a sample survey stratified by
regional council, industry and size. The series starts at 2002 because of a break
in the target population before that year, which is also why the 1994
observation carried in the same table is excluded.

### What was excluded, and why

- **Territorial authority geography** — published only in census years, so no
  annual series exists at that level.
- **Horticulture** — not collected in many of the years in this window.
- **Forestry** — a definitional break in 2022, and now an MPI rather than a
  Stats NZ collection.
- **Farm type** — an ANZSIC96 to ANZSIC06 recode in 2007 falls inside the window.
- **Pre-2002 data** — the population break described above.
- **A choropleth map** — a sorted bar chart answers "which regions contributed
  most" and a map does not, because a map encodes land area rather than
  magnitude. Canterbury, Otago and Southland are also three of the largest
  regions by area, which is precisely the confound a map would introduce.
- **A second data set** — the question is answerable inside one table, and every
  extra source adds provenance to defend without adding evidence.

### Implications for output design

Three observations follow from the checks above, and they concern the shape of
the output rather than the sheep.

Regional council is the only geography available annually; territorial authority
exists only in census years, so any sub-regional time series is structurally
unavailable rather than merely unpublished. Suppression now runs above ten
percent of cells and is concentrated in the same small regions — Auckland,
Nelson, Chatham Islands — so those regional trends are already unusable, and a
suppression-aware grouping that pooled them would preserve more signal than
suppressing them one at a time does. And the alternating census-and-survey design
makes every regional series in this table structurally uneven, which belongs on
the output rather than being left for a user to discover from a reconciliation
residual.

There is also a class of question this data cannot answer at all: anything at
farm level, or any breakdown finer than what is published. The access route for
those questions is approved microdata under the Five Safes framework, which I
have not used here.

## Limitations

- Non-census years are **sample estimates**. I have not recomputed their sampling
  errors and I have applied no weighting of my own. A year-on-year movement in a
  single small region should not be read as real change.
- Suppressed cells are retained as `NA` and counted. Two regions are excluded
  from the 2002–2025 change entirely for that reason.
- The sheep-to-cattle comparisons are **ratios of head counts**. They are not
  claims about stocking rate, feed demand, land use or emissions.
- The target population is GST-registered agricultural businesses. Coverage of
  the smallest farms is partial and not quantifiable from published data.
- Choosing 2002 and 2025 as endpoints is a choice. A 2015–2025 window gives the
  same qualitative answer, with the same three regions accounting for 59 percent
  of the fall rather than 54 percent.
- Input perturbation has been applied to agricultural statistics since 2017. It
  is small relative to every effect discussed here, but it is present.

## Ethics

This analysis uses published regional aggregates only. No microdata and no IDI
access is involved, so the Five Safes framework does not apply to it.

Stats NZ identifies Māori-owned farms in this collection by matching the
Agricultural Production Survey against the Tatauranga umanga Māori population.
Under Māori data sovereignty principles, and under the Mana Ōrite Relationship
Agreement, the authority to interpret and narrate that data sits with iwi Māori
and Stats NZ. It does not sit with an unaffiliated analyst. That is why this
report does not comment on it.

## Reproduce

```bash
quarto render
```

`R/load.R` reads and tidies the pinned extract, `R/checks.R` runs the validation
rules, and `index.qmd` produces this page and writes the three figures into
`outputs/figures/`.

```{r}
#| label: session
#| code-fold: false
sessionInfo()
```
